Single-cell RNA sequencing of the holothurian regenerating intestine reveals the pluripotency of the coelomic epithelium

Version 1

Abstract

Abstract

In holothurians, the regenerative process following evisceration involves the development of a “rudiment” or “anlage” at the injured end of the mesentery. This regenerating anlage plays a pivotal role in the formation of a new intestine. Despite its significance, our understanding of the molecular characteristics inherent to the constituent cells of this structure has remained limited. To address this gap, we employed state-of-the-art scRNA-seq and HCR-FISH analyses to discern the distinct cellular populations associated with the regeneration anlage. Through this approach, we successfully identified thirteen distinct cell clusters. Among these, two clusters exhibit characteristics consistent with putative mesenchymal cells, while another four show features akin to coelomocyte cell populations. The remaining seven cell clusters collectively form a large group encompassing the coelomic epithelium of the regenerating anlage and mesentery. Within this large group of clusters, we recognized previously documented cell populations such as muscle precursors, neuroepithelial cells and actively proliferating cells. Strikingly, our analysis provides data for identifying at least four other cellular populations that we define as the precursor cells of the growing anlage. Consequently, our findings strengthen the hypothesis that the coelomic epithelium of the anlage is a pluripotent tissue that gives rise to diverse cell types of the regenerating intestinal organ. Moreover, our results provide the initial view into the transcriptomic analysis of cell populations responsible for the amazing regenerative capabilities of echinoderms.

Introduction

Animals exhibit a remarkable diversity in their responses to injury, ranging from basic wound healing to the complete regeneration of lost structures. At one end of the spectrum are species that heal wounds without regenerating the missing part, while at the other end are those capable of recreating structures identical to the original. Despite these differences, all animals possess some level of regenerative ability. This capacity for regeneration can vary not only between different species but also within the same species, depending on the tissue or organ involved.

For decades, scientists have tried to understand these vast differences in regeneration capacities across the animal kingdom. For this they have focused on those species that show amazing regenerative abilities (Tanaka and Reddien, 2011), including coelenterates (hydra) (Vogg et al., 2019), flatworms (planaria) (Reddien, 2018), fish (zebrafish) (Gemberling et al., 2013) and amphibians (axolotl) (Roy and Gatien, 2008) among others. These studies have uncovered various mechanisms that regeneration-competent species exhibit to regenerate tissues, entire organs, body parts, and in some cases complete bodies. Key findings include the discovery of essential processes, such as the formation of a blastema – a mass of proliferating cells that plays a crucial role in regenerating the lost structure (Min and Whited, 2023; Poleo et al., 2001; Santos-Ruiz et al., 2002; Seifert and Muneoka, 2018; Stocum, 2004; Zenjari et al., 1996).

Among deuterostomes, echinoderms are considered prime exponents of regenerative capability (Candia-Carnevali et al., 2024). Within this group, holothurians, commonly known as sea cucumbers, exhibit an extraordinary form of regeneration. They can regenerate their internal organs following evisceration, a process in which they expel their viscera in response to stress or predation (Byrne, 2023). This extraordinary ability makes them a valuable model for studying regeneration in complex organisms. The regeneration of the digestive system in holothurians, in particular, has garnered significant interest (Dolmatov, 2021; Mashanov et al., 2014; Medina-Feliciano and García-Arrarás, 2021; Quispe-Parra et al., 2021; Su et al., 2022). Upon evisceration, the holothurian intestine, which constitutes nearly their entire digestive tract, begins to regenerate from the mesentery, a supportive tissue layer where the original intestine was attached (García-Arrarás et al., 1998). A thickening at the injured end of the mesentery, known as an “anlage” or “rudiment,” initiates this process (García-Arrarás et al., 1998). This structure is analogous to a blastema but differs in that the cell proliferation mainly occurs in the surrounding epithelium, rather than in the mesenchymal cells, as observed in classical blastemas (Carlson, 2007; García-Arrarás et al., 1998).

Histologically, the holothurian intestinal anlage forms from dedifferentiated cells within the mesentery, which revert to a more stem-cell-like state before proliferating and migrating to form a new intestinal structure (Candelaria et al., 2006; Mashanov et al., 2005). This dedifferentiation process is crucial for regeneration, involving a spatial and temporal gradient starting at the injury site and extending along the mesenteric border (García-Arrarás et al., 2011). The new epithelial layer that forms around the anlage is distinct from the original mesothelium, shows significant morphological and molecular changes compared to the mesenteric tissue (Candelaria et al., 2006; García-Arrarás et al., 1998; Mashanov et al., 2005). Further examination of holothurian regeneration reveals that most cellular division occurs within the anlage’s coelomic epithelium (García-Arrarás et al., 2011, 1998). These proliferating cells are hypothesized to differentiate into various cell types, including myocytes and neurons, as well as mesenchymal cells, through an epithelial to mesenchymal transition (EMT) (García-Arrarás et al., 2011, 1998). The gene expression profiles during the formation and growth of the anlage suggest extensive reprogramming, leading to a more plastic cell phenotype (Ortiz-Pineda et al., 2009; Quispe-Parra et al., 2021; Rojas-Cartagena et al., 2007).

The process of intestinal regeneration in holothurians raises fundamental questions of regenerative phenomena particularly concerning the identity, origin, and fate of progenitor cells, involved in the process (Alvarado and Tsonis, 2006; Candia-Carnevali et al., 2024). In this context, the role of the anlage, and specially the mesothelium (also named celothelium) in echinoderms, deserved particular attention (Candia-Carnevali et al., 2024; Smiley, 1994). This tissue, mainly composed of coelomic epithelia and myocytes, exhibit significant morphological and gene expression changes that are associated with the dedifferentiation process. These dedifferentiated cells form the coelomic epithelium of the anlage and appear to be the principal source of cells for the new intestine. Despite these findings, little is known about the cell composition and dynamics of the anlage nor of its coelomic epithelium.

Single-cell omic studies offer tremendous promise to dissect the cellular contributions of the holothurian intestinal anlage and to identify the specific cells involved in generating a new intestine. Single-cell RNA sequencing (scRNA-seq) a powerful tool for dissecting cellular composition and dynamics has been used in related species to explore regenerating or developing tissues. In mice, for instance, scRNA-seq has been employed to investigate the response to injury. However, these studies have primarily focused on tissue healing rather than the full regenerative process (Ayyaz et al., 2019; Beumer and Clevers, 2021; Capdevila et al., 2021; Parigi et al., 2022). Beyond healing, scRNA-seq has been pivotal in studying regenerating systems such as the axolotl limb blastema and planarian regeneration providing a detailed insights into the cellular processes at play (Gerber et al., 2018; King et al., 2024; Leigh et al., 2018; Rodgers et al., 2020). Furthermore, considering the close relationship between regeneration and embryogenesis, scRNA-seq, has also been used to understand the cellular events in developing echinoderm embryos (Cocurullo et al., 2023; Paganos et al., 2022a, 2022b, 2021; Satoh et al., 2022; Tominaga et al., 2023). These applications highlight the groundbreaking role of scRNA-seq in advancing our knowledge of the cellular mechanisms in both regenerative and developmental contexts.

In this study, we employ scRNA-seq to analyze the regenerating intestinal anlage of the sea cucumber Holothuria glaberrima, aiming to delineate its constituent cellular populations (Fig 1A). We corroborate our findings using hybridization chain reaction fluorescent in situ hybridization (HCR-FISH) to verify the presence and location of specific cell types (Choi et al., 2018). The results provides transcriptomic evidence on the pluripotency of the anlage coelomic epithelia and the first description of the holothurian regenerating precursor cells. This research not only advances our understanding of the unique regenerative capabilities of holothurians but also contributes to the broader field of regenerative biology, highlighting the diverse strategies employed by different organisms to restore lost tissues.

Overview of single cell RNA sequencing of regenerating intestinal tissue of H. glaberrima.

(A) Diagram of 9-day regenerating intestine depicting mesentery and anlage components (B) UMAP plot of population identities determined through unsupervised clustering of 9-day regenerating mesentery and anlage tissues. (C) Representation of cells per tissue of both samples. (C) UMAP projections of cluster cells separated by tissue of origin. (D) Percentage of cells per cluster based on their tissue of origin.

Results

Previous work from our laboratory has shown that the rudiment or anlage that forms at the tip of the mesentery plays a pivotal role in the formation of the new intestine (García-Arrarás et al., 2019). This transient mass of cells is thought to give rise to most intestinal cell types, the sole exception being the luminal cells. Therefore, to maximize the characterization of the cells in the regenerative anlage, we chose to perform scRNA-seq in the tissues of 9-days post evisceration (dpe) regenerating animals (García-Arrarás et al., 2011, 1998). At this stage a well-formed anlage consists of epithelial cells surrounding a large area of connective tissue populated with mesenchymal cells (Fig 1A). More importantly, different cell populations undergoing proliferation or differentiation can be found at this stage. Since some of the cellular processes during regeneration occur in a spatio-temporal gradient along the length of the mesentery, we separately surveyed the anlage and the mesentery tissues accordingly. Thus, for each animal, the anlage was separated from the mesentery, and both tissues were processed independently, for a total of four scRNA-seq runs.

Tissues were treated with 0.05% trypsin for 15 min and cells were dissociated manually. The dissociated cells were counted and used for the scRNA-seq analyses. Additionally, samples from the cell dissociation were fixed in poly-lysine treated slides and analyzed with immuno- and cyto-chemical techniques. Therefore, these cells correspond to the same batch whose transcriptomes were sequenced.

Cell Heterogeneity: Immuno- and cyto-chemical analyses

The strength of the scRNA-seq data depends mainly on the dissociation and isolation of the cell populations from the dissected tissue. Since our focus was on the cells of the regenerating anlage, we devised a dissociation protocol that favored the isolation of the cells within this structure. To determine, at least partially, the cell types in our original dissociation, we performed immuno- and cyto-chemical analyses on the enzyme-dissociated cell suspension that was used for the scRNA-seq. The labeling obtained for each marker is shown in S1 Fig.

We found that, in the dissociated anlage, 5% of the cells were labeled with a spherulocyte (immune cell) marker, 2-7% with phalloidin (a muscle marker), 40% with a mesenchymal marker (KL14-antibody), and a large number of cells (32-60%) with a mesothelial marker (MESO-antibody). Similar populations were found in the mesentery, although in this tissue, 5-10% of cells expressed the neuronal markers, heptapeptide GFSKLYFamide (Diaz-Miranda et al., 1995) and RN1 (Diaz-Balzac et al., 2007). This analysis suggested that most of the cells originate from the regenerate coelomic epithlium.

Different populations were also observed with three different tubulin antibodies (S2 Fig). In the 9-dpe anlage, anti-acetylated alpha tubulin labeled about 7% of the cells, anti-beta tubulin labeled about 70% of the cells, while an anti-alpha tubulin labeled about 80% of the cells. Except for the acetylated alpha tubulin, which labeled about 1% of the cells dissociated from the mesentery versus 7% of those from the anlage, anti-alpha and beta labeling percentages were similar in both mesentery and anlage dissociated cells.

Finally, a cell population labeled with fluorescently-labeled phalloidin accounted for about 7% of the cells in both mesentery and anlage. These cells, however, did not correspond to the elongated muscle cells of the mesentery. Instead, they were rounded cells with labeling found in the cytoplasm surrounding one side of the nuclei.

Two cell populations from the mesentery are not represented or are greatly underrepresented in the scRNA-seq. Firstly, the muscle cells of the mesentery, due to their non-dissociation by the protocol used, were absent in the dissociated cell suspension. Their elongated morphology would have further complicated their passage via cell separation system for sequencing. Secondly, the majority of neurons from the neuronal network associated with the mesentery could not be dissociated (Nieves-Ríos et al., 2020). We did, however, observe structures that resembled a tangled mass of cells immunoreactive to some of our neuronal markers. This suggests that the mesentery nervous component, being unable to be isolated as single cells, was not sequenced.

In summary, the immuno- and cytochemical results show that the dissociated cell populations sequenced correspond to cellular phenotypes that have been previously described within the regenerating anlage. The abundance of these cells in the sequenced samples corresponds to the ease of their dissociation by trypsin. Thus, dedifferentiated cells of the mesentery and anlage epithelium (which are loosely connected to each other) and those of the connective tissue are probably over-represented compared to differentiated cell types.

Cell populations defined by scRNA-seq

Analysis of scRNA-seq data resulted in a total of 3,844 cells, with 2,392 originating from the two anlage samples and 1,452 from the two mesentery samples. Upon dataset integration and graph-based clustering, we identified 13 clusters, each thought to represent singular cell types or cell states in the regenerating intestine (Fig 1A-B). The percentage of cells that form each cluster differs from cluster to cluster, ranging from 21% (cluster 0) to 1% (cluster 12) of the total cells. Nonetheless, each cluster consists of cells from both the mesentery and anlage samples (Fig. 1C & D). The number of clusters did not change dramatically under various parameters (resolution, dimensionality, and number of variable features), constantly around 12-15 clusters. Moreover, except for C3 and C4, the clusters are supported by the clustering significance analysis performed with scSHC (Grabski et al., 2023), a model-based hypothesis testing method for scRNA-seq that evaluates the probability of each individual cluster being unique (S3 Fig). The uniqueness of C3 and C4 is suggested by additional analyses as will be addressed below.

Each identified cluster exhibits a distinctive gene expression profile relative to cells in other clusters (Fig 2). The top expressed gene serves as a marker for each cell population to highlight the uniqueness of each cluster. In Fig 2, we show the relative expression of the top gene for each cluster based on two factors: (1) difference in percentage of representation and (2) log2 fold-change (log2FC) against all other clusters. Interestingly, each cluster shows dramatic differential expression values (> 2 log2FC) and differences in representation percentages over 50%, except for C0 through C3, with differences in representation around 30%.

UMAP visualization of clusters highlighting the expression of their top genes.

Each gene corresponds to the top gene of each independent cluster based on the percentage of representation to other clusters.

Prior to characterizing each of the 13 cell clusters, we sought to understand what, in a broad view, appears to be a segregation of ∼90% of the cells into two distinct supra-clusters. One of them encompassing 7 clusters (C0, C1, C3, C4, C5, C8, C9) that corresponded to 69.6% of all cells and the other encompassing two clusters (C2 and C7) that corresponded to 19.1% of cells. The remaining 11.3% of cells were distributed in 4 distinct isolated clusters (C6, C10, C11, and C12). As stated earlier, all clusters have representation from mesentery and anlage tissues (Fig 1C-D), thus excluding the possibility that the two supra-clusters represented mesentery versus anlage cells.

Mesenchymal versus Epithelial Clusters

These two supra-clusters are of interest as they appear to represent the two main cell types found in the regenerating intestine: coelomic epithelial cells and mesenchymal cells (Fig 1B). Understanding the cell types that correspond to each cluster is essential as cellular processes crucial for regeneration are thought to be localized to the coelomic epithelia of the regenerating anlage. Comparison between these two supra-clusters showed distinct expression profiles that allowed us to characterize their cell types. For instance, the supra-cluster composed of C2 and C7, showed ERG (transcriptional regulator ETS-related gene) as the top expressed gene, an oncogene that is related to embryonic development, differentiation, angiogenesis, and apoptosis (Dhordain et al., 1995; Iwamoto et al., 2001; Vlaeminck-Guillem et al., 2000) (Fig 2). More importantly, the expression of ERG has been associated with mesenchymal identity in other echinoderms (Meyer et al., 2023; Tominaga et al., 2023). In these studies, ERG has been reported as the marker gene of mesenchymal cells of sea urchin and sea star larva by scRNA-seq analyses and the localization of ERG on embryonic precursor mesenchymal cells of the sea urchin was further confirmed by in situ analyses (Meyer et al., 2023). In addition, this latter group also reported these mesenchymal cells to express a GATA transcription factor (GATA3) and ETS1, which in our dataset are also being expressed only by populations within this supra-cluster (GATA2 and ETS1 – Fig 3B). The gene ETS1 has been associated with the differentiating sea urchin larvae mesenchymal cells and has a vital role in EMT and cellular invasion of mesenchymal cancer cells (Gluck et al., 2019; Koga et al., 2010; Meyer et al., 2023). The expression of PRG4 (proteoglycan-4) reinforces the mesenchymal identity of this supra-cluster in the regenerating intestine (S4 Fig), which correlates with previous studies that have identified a proteoglycan-like molecule in the mesenchyme of 7-dpe regenerating intestine (Vázquez-Vélez et al., 2016).

Cluster characterization by gene expression.

(A) Top-expressed genes or (B) genes corresponding to transcription factors and intercellular signaling molecules are identified in the 13 cell clusters. Clusters are classified by their corresponding cell type, where blue corresponds to cells of the coelomic epithelium, red to those in the mesenchyme and green to coelomocytes. Color intensity shows the expression level of each gene in log2fold-change (log2FC) values. Dot size corresponds to percentage of representation of the gene in the respective cluster compared to all others. Gene identifiers starting with “g” correspond to uncharacterized gene models of H. glaberrima. (C) UMAP plot of clusters colored by cell type.

In contrast, the top marker gene in the other supra-cluster is AHNK, known as neuroblast differentiation-associated protein AHNAK. Reports have shown AHNK to have a role in calcium regulation, cellular migration, and carcinogenic transformation of colon epithelial cells (Dumitru et al., 2013). Furthermore, this gene is overexpressed in regenerating rat muscle compared to normal muscle (Huang et al., 2007). However, the localized expression of WNT9 in the C1, C3, C5, and C8 of this supra-cluster more clearly favors its classification as a marker for coelomic epithelial cell types (Fig 3B and S4 Fig). A previous study from our laboratory using in situ hybridization, details the expression of WNT9 during intestinal regeneration in H. glaberrima, where it was shown to be localized to the coelomic epithelium of the anlage and adjacent mesentery (Mashanov et al., 2012). In addition, correlating with what was shown by the in situ hybridization results, the population of cells that differentially expresses WNT9 makes 15% of those in the mesentery. However, it is close to 30% of the cells in the anlage. Additional analyses, discussed below, further strengthens the coelomic epithelium identity of cells in this supra-cluster.

In summary, results show 13 individual cell clusters with distinct expression profiles in the regenerative intestinal tissue. Moreover, our results suggest a principal separation of clusters based on whether they correspond to those of the mesenchyme or to cells from the coelomic epithelia layer. The rest of the populations show top expressed genes that are immune-related suggesting that these must be coelomocyte populations (Fig 2).

Cluster Identities

Rather than considering single genes, the uniqueness of each cluster can be assessed in terms of their transcriptomic profile (Fig 3A) and by the expression of transcription factors and intercellular signaling molecules (Fig 3B). As shown in Fig 3A-B, each cluster has a unique transcriptomic profile. Nonetheless, some clusters show their close relationship (epithelial vs mesenchymal) based on shared gene expression or representation. A notable example is clusters C0, C1, C3, and C4 of the epithelial layer populations, which share many top genes, albeit at different expression levels. Similarly, C8 and C9 have low expression of genes expressed by C0, C1, C3 and C4 but have expression of other genes that are not expressed by any other cluster. This is also true for the mesenchymal clusters, where these two clusters have overlap of some genes, that are not expressed by other clusters. Therefore, these results show the interaction of these clusters and their transcriptomic relationship depending on where they are localized within the regenerating intestinal tissue.

Identifying the top genes expressed by each cluster also provides essential, notwithstanding limited, information for their complete identification. To further characterize each cluster, we have used other analyses or performed additional experiments, including (1) multiple gene expression patterns, (2) enriched ontology, (3) HCR-FISH and (4) pseudo-trajectory. Initially, we describe in depth the potential identity of these clusters based on their expression pattern and enriched ontology, and we will end with their potential interactions.

An in-depth analysis of the various clusters and their possible relation to cell populations

Coelomocyte populations

Coelomocytes are specialized cells found in the coelomic fluid and within organs of echinoderms. These cells have been associated with immunological roles including pathogen recognition, encapsulation, phagocytosis, debris removal, cytokine production and secretion among others (Barela Hudgell et al., 2022; Courtney Smith et al., 2018; García-Arrarás et al., 2006; Smith et al., 1995), In holothurians, these cells have been shown to be present at injury sites and in the regeneration anlage. Coelomocytes can be subdivided into different populations by using morphological, physiological, and molecular characteristics (Ramírez-Gómez et al., 2010).

Many of the coelomocyte characteristics correlate with the top differentially expressed genes of cell populations within our data. These immune-like clusters (C6, C10, C11, and C12) have high expression of genes related to the immune system that are not shared with any other cluster. For example, C6 embodies a distinct cell population that represents a substantial number of cells (6% of all cells, 8% of the mesentery and 4% of the cells in the anlage). These cells are the only population expressing FBCD1 (fibrinogen C domain-containing protein 1). Cells in C6 also express other genes such as various tyrosine protein phosphate receptors, integrin alpha-8 (ITA8), platelet glycoprotein V (GPV), leucine rich repeats and immunoglobulin-like domains protein 2 (LRIG2) (Fig 4A). The immune identity of C6 is also supported by the resulting gene set enrichment of gene ontology (gseGO) terms related to immune responses, such as ubiquitin-dependent ERAD pathway, innate immune response activating cell surface receptor signaling pathway, respond to endoplasmic reticulum stress, phagocytosis and many more related to defense mechanisms (Fig 4B). Similarly, C10 and C12 transcriptomic profiles suggest they correspond to immune-like populations (Fig 4C). Specifically, C10 shows various uncharacterized genes along with HCK (tyrosine-protein kinase HCK), DMBT1 (deleted in malignant brain tumors 1), ALS (insulin-like growth factor binding protein complex acid labile subunit), GPV, and IRF8 (interferon regulatory factor 8), FER (tyrosine-protein kinase Fer). The top GO enriched terms of this cluster support its involvement in immune process, some of these being: lipase activity regulation, Fc receptor mediated stimulatory signaling pathway, cellular pigmentation, B cell activation involved in immune response, and Fc receptor signaling pathway. The transcripts expressed by C12 show a more complex profile, where most of the top genes are uncharacterized. Nonetheless, among the annotated genes are BAR3 (Balbiani ring protein 3), LYS1 (lysozyme 1), TRPM3 (transient receptor cation channel subfamily M member 3), PGCA (aggrecan core protein) and MRC1 (macrophage mannose receptor 1). The top genes of C11 include a great number of immune genes such as MUC5A (mucin-5AC), FCGBP (IgGFc-binding protein), SSPO (SCO-spondin), FCN1A (ficolin-1-A), MUC5B (mucin-5B), and TIE1 (tyrosine-protein kinase receptor Tie-1). However, the top genes of this cluster also include several genes involved in neuronal activity, which is evident in the top enriched GO terms of this cluster that include regulation of postsynaptic membrane potential, excitatory postsynaptic potential, chemical synaptic transmission, endoplasmic reticulum to Golgi vesicle-mediated transport, adult behavior, and regulation of neurotransmitter levels (Fig 4A-B).

Expression profile of coelomocyte cell types.

(A) Dot plot showing the top expressed genes of C6, C10, C11, and C12. Color intensity reflects the expression levels in log2FC values. Dot size correspond to the percentage of representation of each gene in the respective cluster compared to all others. (B) Top enriched GO biological processes terms of C6, C10, C11, and C12. Dot size corresponds to the enrichment score while the color reflects the adjusted p-value. (C) UMAP highlighting the coelomocyte populations. (D-E) HCR-FISH for FBCD1 in holothurian intestinal tissues. Cells expressing FBCD1 mRNA in the connective tissue layer of (D) regenerating mesentery and (E) normal intestine of H. glaberrima are shown. Red = DAPI stain, green = HCR-FISH. Bar= 10 μm.

To partially confirm the prediction that cells from C6, C10, C11 and C12 corresponded to coelomocyte populations, we used HCR-FISH to identify the cell types expressing the top gene in one of the clusters. We focused on the expression of FBCD1, the gene that is the most represented by cells of C6. In situ hybridization identified a distinct cell type in both regenerating and non-regenerating tissues (Fig 4D-E). These cells showed round or oval morphologies with a central round nucleus. In some cases, short extensions could be observed. The cells were heterogeneously distributed in all tissues, including the nervous system and the body wall, and could be found associated with either epithelial tissues or with the extracellular matrix (ECM) in the normal intestine, the intestinal anlage and in the mesentery of normal or regenerating animals. The widespread distribution of this cell type hinted at a cell function consistent with patrolling the body to detect and respond to potential threats such as injury or bacterial invasion.

Mesenchymal cell populations

Mesenchymal cells of the intestinal anlage are yet to be well studied. They are known to be less proliferative than those in the overlying epithelium (García-Arrarás et al., 2011, 1998) and involved in ECM remodeling (Quiñones et al., 2002). Some of the mesenchymal cells are thought to migrate from the connective tissue of the mesentery (Cabrera-Serrano and García-Arrarás, 2004), while others have been shown to originate via EMT from the overlying epithelium (García-Arrarás et al., 2011). Some of these cells will form a mesenchymal cellular layer associated with the luminal epithelial cells as the lumen forms (García-Arrarás et al., 2011). As explained earlier, we have proposed that C2 and C7, form a separate supra-cluster, clearly identified by the increased expression of ERG with 5 of log2FC and a 90% representation (Fig 2O). These two clusters also share many marker genes, including multiple ECM genes (Fig 5D). For instance, highest expressed gene of C2 is TIMP3 (metalloproteinase inhibitor 3), followed by NPNT (nephronectin), which has been reported to be an integrin ligand during kidney development (Sun et al., 2018). Additionally, this cluster has overexpression of ECM2 (extracellular matrix protein 2), KLKB1 (plasma kallikrein), MMP14 (matrix metalloproteinase 14), MMP24 (matrix metalloproteinase 24), HMCN1 (hemicentin-1), and ITA8 (integrin alpha-8), all of which are explicitly related to extracellular matrix component (Bokel and Brown, 2002; Dolmatov and Nizhnichenko, 2023; Dong et al., 2006; Stamenkovic, 2003; Volkert et al., 2014). Many of these genes are also highly represented in C7. However, here we also find ITIH2 (inter-alpha-trypsin inhibitor heavy chain H2), DUOX1 (dual oxidase 1), SVEP1 (sushi, von Willebrand factor type A, EGF and pentraxin domain-containing protein 1), SEPP1 (selenoprotein P), and KLH20 (Kelch-like protein 20) suggesting that cells have gained some specialization and are more advanced in their differentiation when compared to those of C2. Along with this, when analyzing the gseGO, for C2 we obtain GO terms of numerous ECM processes, such as cell adhesion mediated by integrin, integrin-mediated signaling pathway, and regulation of cell-substrate junction assembly (Fig 5F). Similarly, C7 has an enrichment of regulation of cell-matrix adhesion, regulation of cell-substrate junction organization, and substrate adhesion-dependent cell spreading. Notably, these clusters also show enrichment of processes involved in cell proliferation, growth, and wound healing. To mention a few for C2 we find B cell differentiation, vascular endothelial growth factor receptor signaling, regulation of myeloid leukocyte differentiation, and negative regulation of myeloid cell differentiation, while C7 shows an enrichment of positive regulation of wound healing, regulation of neural precursor cell proliferation, positive regulation of response to wounding, humoral immune response, neuroblast proliferation, among others. This is not surprising considering that this tissue is undergoing active regeneration which requires growth and healing events to be active at distinct stages of the process.

Expression profile of mesenchymal cell types.

(A) Dot plot showing the top expressed genes of C2 and C7. Color intensity reflects expression levels in log2FC values. Dot size correspond to the percentage of representation of each gene in the respective cluster compared to all others. (B) Top GO enriched terms of biological processes of C2 and C7. Dot size corresponds to the enrichment score while the color reflects the adjusted p-value. (C) UMAP highlighting the clusters that correspond to the mesenchymal population (D-F) Cells expressing HMCNT1 mRNA in the connective tissue layer of (D) regenerating mesentery, (E) intestinal anlage and (F) normal intestine of H. glaberrima. Red = DAPI stain, green = HCR-FISH. Bar= 10μm

To confirm our prediction that cells in C2 and C7 were those present within the mesenchyme, we chose to localize the expression of HMCNT1 mRNA. The mRNA for this protein, known to code for an extracellular protein (Dolmatov and Nizhnichenko, 2023; Lindsay-Mosher et al., 2020; Welcker et al., 2021), is present in both cell clusters. HCR-FISH showed that cells expressing the HMCNT1 mRNA were present in the connective tissue of the regenerating mesentery and certain regions of the connective tissue of the anlage. In the normal intestine, labeled cells were present in the connective tissue (Fig 5A-C). Cells in the mesentery and normal intestine connective tissue were somewhat distanced from each other and had an irregular morphology with intense labeling throughout the cytoplasm. A weaker labeling was observed in cells of the anlage, which were more densely packed and adjacent to the coelomic epithelium, suggesting that they correspond to cells undergoing EMT on their way to differentiate into ECM-producing mesenchymal cells. A very similar pattern of expression was observed with HCR-FISH for ERG (not shown). The ERG-expressing cells were also found in the connective tissue of normal intestine and in the cells of the anlage undergoing EMT.

Coelomic epithelia/mesothelial cell populations

The mesothelial layer of the intestine and mesentery is formed by cells in contact with the coelomic fluid (coelomic epithelium or peritoneocytes) together with myocytes and neurons. In the regenerating tissues, some of these cells dedifferentiate and form a coelomic epithelia that differs in morphology (Mashanov et al., 2005) and gene expression (Mashanov et al., 2017, 2015, 2012, 2010) to the mesothelium that normally surrounds the organ. This dedifferentiated coelomic epithelium which is present mainly in the anlage and in areas of the adjacent mesentery, is also responsible for most of the cell division that takes place in the regenerating intestine (García-Arrarás et al., 2011). Many of the changes observed in the mesothelium during regeneration take place in a gradient, beginning at the tip of the mesentery (where the anlage forms) and continuing in the adjacent regions of the mesentery mesothelium. Thus, the analysis of the scRNA-seq data in view of our knowledge of the ongoing events in the 9-dpe regenerating organ leads to the conclusion, as stated previously, that the 7 clusters within the major supra-cluster, represent the cells in the coelomic epithelium of the mesentery and the anlage. Here is our analysis:

C8 represents the proliferating cells

These cells mainly found within the anlage coelomic epithelia express proliferation markers such as PLK1 (serine/threonine-protein kinase PLK1), SMC2 (structural maintenance of chromosomes protein 2), PRI2 (PRIM2 - DNA primase large subunit), PCNA (proliferating cell nuclear antigen), and CDK1 (cyclin-dependent kinase 1) (Locard-Paulet et al., 2022). In addition, this cluster has high expression of TOP2A (DNA topoisomerase 2-beta), a gene that has also been seen to be overexpressed in proliferating basal cells of the human gastrointestinal epithelia (Busslinger et al., 2021). Other marker genes related to cell mitotic activity found here include CENPE (centromere-associated protein E), SMC4 (structural maintenance of chromosomes protein 4), K167 (proliferation marker protein Ki-67), and CCNB3 (G2/mitotic-specific cyclin-B3) (Fig 6A). Furthermore, this cluster expresses specific transcription factors such as E2F3, MCM10 (protein MCM10 homolog), PAF15 (PCNA-associated factor) and BRCA1 (breast cancer type 1 susceptibility protein) that are also associated with control of cell division (Humbert et al., 2000; Lõoke et al., 2017; Xie et al., 2014). The gseGO terms also confirm its proliferative identity with GO terms related to chromosome separation, condensation, mitotic cytokinesis, and regulation of cell cycle checkpoint (Fig 6B). Moreover, the dividing cell population is higher in the anlage samples (4%) than in the mesentery samples (2%), which is in accordance with what we have observed in regenerating animals, that while cell division does take place in the mesentery, more cells are proliferating in the epithelial layer of the intestinal anlage (Bello et al., 2020; García-Arrarás et al., 2011, 1998).

Expression profile of cell clusters undergoing differentiation or proliferation in the coelomic epithelium.

(A) Dot plot showing the expression profile of C5, C8, and C9. Color intensity reflects expression levels in log2FC values. Dot size represents percentage of representation of each gene. (B) Top GO enriched terms of biological processes of each cluster. (C) UMAP highlighting C5, C8, and C9, which along with clusters in color blue correspond to the coelomic epithelium. (D-F) Cells expressing MYH7 mRNA in the (D) regenerating mesentery, (E) basal area of the anlage coelomic epithelium and (F) muscle layer of the normal intestine. (G) Phalloidin labeled differentiating muscle cells in basal area of the coelomic epithelium. (H) Cell labeled with RN1 antibody (neuronal marker) in coelomic epithelia of the anlage. Red = DAPI stain, green D-F= HCR-FISH. G= Fluorescent Phalloidin, H= RN1 antibody. Bar-D-F= 25μm, G-H=10μm.

C5 represents muscle precursors

A second cell population known to originate from the coelomic epithelium and that can be identified in our data is that of muscle precursors. These cells are known to differentiate into enteric muscle during the second week of regeneration (Murray and García-Arrarás, 2004). This population can be recognized in our data by the expression of muscle-specific markers present in C5, such as TITIN, MYL1 (myosin light chain 1/3), MYH7 (myosin 7), ACTG (actin, cytoplasmic 2), and CNN3 (calponin-3) (Fig 6A). Considering that this cell population is potentially a population of the coelomic epithelium actively undergoing differentiation toward muscle phenotype, they also share gene expression with other epithelial cell clusters (C0, C1, C3, C4), albeit at a lower fold-change. Similarly, this cluster shows a specific expression of transcription factors associated to muscle cells, namely FXL16 (F-box/LRR-repeat protein 16) and SCRT2 (transcriptional repressor scratch 2). Results of GO terms of this cluster also demonstrate enriched terms related to muscle tissue growth, such as muscle tissue morphogenesis, muscle development, myofibril assembly, and sarcomere organization (Fig 6B). The focus on development and morphogenesis is to be expected, considering that these are still undergoing differentiation toward a muscle phenotype. HCR-FISH of MYH7 corroborates the muscle precursor phenotype of the cells in this cluster (Fig 6D-F). Labeling is observed in the same regions as myoblasts or muscle cells were previously identified (Murray and García-Arrarás, 2004), including differentiating muscle cells within or underlying the coelomic epithelia of the mesentery and the anlage (Fig 6D & 6E) and, as expected, the muscle layer in the normal intestine (Fig 6F). The same cell population found in the basal side of the coelomic epithelia of the regenerating intestine is labeled with the muscle marker, fluorescently-labeled phalloidin as shown in Fig 6G.

C9 represents the neuroepithelial cells

Cells in C9, represent another population of specialized cells that can be associated to cells previously described in the intestinal anlage. This small number (3%) of cells most likely corresponds to neuroepithelial cells that will eventually give rise to neurons. These cells are shown to be expressing neuroepithelial or neuronal genes such as neurotrypsin (NETR), potassium gated-voltage channels, PRD10 (PRDM10 - PR domain zinc finger protein 10), ELAV2 and STA10 (STARD10 - START domain-containing protein 10). The latter is a protein that we have characterized as being expressed by enteric neurons and nerve bundles (Rosado-Olivieri et al., 2017) (Fig 6A). The holothurian STA10 (STARD10) is recognized by our monoclonal antibody RN1. This antibody has been used to detect enteric neurons as they begin to differentiate in the coelomic epithelium during the second week of regeneration (Tossas et al., 2014). Labeling of the 8-dpe regenerating mesentery shows one of these RN1 neuronal cells in the coelomic epithelium of the regenerating anlage (Fig 6H). Among its associated GO terms are positive regulation of ion transmembrane transporter activity, cyclic nucleotide metabolic process, regulation of muscle contraction, regulation of membrane potential and positive regulation of hormone secretion (Fig 6B). Moreover, there is also a great representation of processes involved in development, differentiation, and growth of nerve cells, all of which together would be expected of a neuroepithelial layer.

The characterization of these 3 clusters, that represent cells undergoing differentiation or proliferation, leaves a group of 4 clusters (C0, C1, C3, and C4) that show some overlap in expressed genes and at the same time share some gene expression with some of the previously described clusters. Nonetheless, as seen in Fig 3A&B, the cells in these four clusters still have high expression and representation of specific transcripts.

C4 represents the intestinal coelomic epithelial cells

The gene expression profile of cells in C4 sets them slightly apart from the other three clusters (C0, C1, C3). It identifies cells that are more advanced in their development toward a particular phenotype. C4 shows high expression of genes such KCNQ5 (potassium voltage-gated channel subfamily KQT member 5), SC6A9 (sodium and chloride dependent glycine transporter 1), EFNB2 (Ephrin-B2), and UNC5C (Netrin receptor UNC5C) (Fig 7A). Other than these, C4 also shows high expression of genes that are related to cell-cell interactions and ECM molecules such as LAMA2 (laminin subunit alpha-2), MEGF6 (multiple epidermal growth factor-like domains protein 6), FMN1 (formin-1), NPHN (nephrin). This cluster, distinct from others, shows enrichment of GO terms related to more advanced stages of development, such as morphogenesis of epithelium, sensory perception, regulation of calcium ion transmembrane transport, among others (Fig 7B). Noteworthy, cells from this cluster correspond mostly to cell from the mesentery samples (67%) rather than from the anlage (37%). Additionally, from all the mesentery cells, about 11% are part of this cluster, while only 4% of the anlage cells are represented here. HCR-FISH of SC6A5 (sodium- and chloride-dependent glycine transporter 2), a gene differentially expressed in the cells of this cluster provided a surprising result. The expression of this mRNA was observed in a few cells in the coelomic epithelia of the regenerating mesentery and intestinal anlage (Fig 7D-E). However, intense labeling is observed in some cells of the normal intestine mesothelium. This pattern of labeling strongly suggests that the labeled cells of the regenerating intestine grouped in C4 (both mesenteryal and anlage) correspond to cells that are in a differentiation pathway to become part of the coelomic epithelia (possibly the peritoneocytes) in the regenerated organ. This conclusion is strengthened by the pseudotime analyses presented in the following section.

Expression profile of coelomic cell populations in the regenerating intestine anlage and mesentery.

(A) Dot plot showing the top expressed genes of C0, C1, C3 and C4. Color intensity reflects expression levels in log2FC values. Dot size correspond to the percentage of representation of each gene in the respective cluster compared to all others. (B) Top GO enriched terms of biological processes of each cluster. Dot size corresponds to the enrichment score while the color reflects the adjusted p-value. (C) UMAP highlighting the clusters of the coelomic epithelial cells of the regenerating intestine. (D-E) HCR-FISH for SC6A5 mRNA was performed as a marker for cells of C4. Labeling is observed in the (D) regenerating mesentery, (E) the coelomic epithelium of the anlage and (F) coelomic epithelium of the normal intestine. Red = DAPI stain, green = HCR-FISH. Bar= 25μm.

C0, C1, and C3 represent differentiation stages of coelomic epithelial cells

The three remaining clusters to be analyzed are C0, C1, and C3. These three clusters share many of their top representative genes, which are associated with developmental, regenerative, or oncogenic processes (Bradford et al., 2009; Dunn et al., 2006; Lu et al., 2015; Nishimoto and Nishida, 2007; Oike et al., 2004; Zhao et al., 2008). These include, for cluster 0: TRFM (melanotransferrin), TIMP4 (metalloproteinase inhibitor 4), and DMBT1, for cluster 1: FGF13 (fibroblast growth factor 13), TGFB3 (transforming growth factor beta 3), HS90A (heat shock protein HSP 90-alpha), and ANGL1 (angiopoietin-related protein 1) (Fig 7A), and for cluster 3: SEM5B (semaphoring-5B), LRIG3 (leucine-rich repeats and immunoglobulin-like domains protein 3), TUTLB (protein turtle homolog B), and NET1 (netrin-1). Moreover, C0 and C1 share an over-representation of ribosomal genes (not shown), and this is reflected in their GO analyses that highlights biological processes related to ribosomal activity such as cytoplasmic translation, and ribosome assembly, biogenesis, and assembly (Fig 7B). In addition to these, C3 also showed enrichment of processes involved in the development of tubular lumen-containing structures such as mesonephric and ureteric ducts, differentiation and regulation of cell growth, and negative regulation of axogenesis (Fig 7B).

C1 and C3 are also closely related in their localization, being overwhelmingly associated with the anlage. These two clusters are mostly composed of cells from the anlage tissue, where we expect to see the precursor cells that will give rise to specialized cells of the organ (Fig 1D). Precisely, C1 and C3 cells together correspond to 37% of all anlage cells, compared to 14.6% of all mesentery cells. In contrast, cells from C0 correspond to 27% and 18% of the mesentery and anlage cells, respectively.

To obtain insight into these cell clusters, we performed HCR-FISH for two different mRNAs; NET1, a chemotropic protein highly represented in C1 and C3, and TRFM, the top represented gene in C0. Both in situ hybridization experiments labeled cells in the coelomic epithelium of the regenerating intestine, supporting our contention that the large supra-cluster represents the coelomic epithelium layer. However, their spatial pattern of expression was unpredictably different. While NET1 was highly expressed in most of the coelomic epithelial cells of the anlage, little expression was found in the regenerating mesentery or in the coelomic epithelium of the normal intestine (Fig 8G-H). TRFM, in contrast, was highly expressed in the coelomic epithelium of the normal intestine and poorly expressed in the intestinal anlage (Fig 8A-E). In the regenerating mesentery, a gradient in expression of the TRFM is observed, where high levels of expression were found in the coelomic epithelium close to the body wall and diminished as one approached the anlage. Thus, the HCR-FISH results show that C0, C1, and C3 correspond to cells of the coelomic epithelium, but strongly suggest that C0 differs from C1 and C3 both in their gene expression profile and in the localization where they are found, both in the regenerating and in the normal intestine.

HCR-FISH for TRFM and NET1 shows the differential gene expression of the anlage versus mesentery coelomic epithelia.

TRFM is mainly expressed in the mesentery coelomic epithelial cells that are distant from the regenerating anlage while NET1 is mainly expressed in the coelomic epithelial cells of the anlage TRFM mRNA expression (A-E) shows a gradient from (A) very strong expression in the coelomic epithelium at the attachment of the mesentery to the body wall, (B) high expression in the coelomic epithelia of the mesentery close to the body wall, (C) weaker expression in the mesentery close to the anlage to (D) weak or lack of expression in the coelomic epithelium of the anlage. (E) TRFM mRNA is expressed by the coelomic epithelium of the normal intestine. regenerating intestine. In contrast, few cells express (A) NET1 mRNA in the regenerating mesentery, while high expression is found in (B) the coelomic epithelium of the anlage (C) Little expression of NET1 is present in the coelomic epithelium of the normal intestine. Red = DAPI stain, green = HCR-FISH. Bar= 25μm.

Finally, it is essential to highlight the many signaling, or growth factors expressed by the coelomic epithelial clusters, in particular C0, C1 and C3 (Fig 3). These include: Wnt, Hox, semaphorin, FGFs, TGF-beta, netrin, insulin-like growth factor (IGF), growth/differentiation factors (GDF), and angiopoietin related proteins (ANGL) (e.g. WNT9, SEM5B, FGF13, NKx3.2, TGFB3, HOX9, IGF1, GDF8, ANGL1, and FOXF1) (Fig 3). This is important in view that the epithelium of the vertebrate blastema is characterized by its chemical modulation of the underlying mesenchyme, as will be discussed later. Likewise, other genes that serve as markers of specific cell types or cellular stages were also identified, including PIWL1 (piwi-like protein 1) in C1, YAP1 in C3 and C4, HES1 (transcription factor HES-1) in C1 and C3, and PRRX1 (paired mesoderm homeobox protein) in C0 and C4 (Fig 3 and S5 Fig).

In summary, of the 13 clusters identified, our data strongly suggest that four of them (C9, C10, C11, and C12) correspond to coelomocytes or immune cells, two of them (C2 and C7) correspond to cells with a mesenchymal phenotype and the remaining 7 to cells of the coelomic epithelia. Of these seven, C5 corresponds to differentiating muscle, C9 to differentiating neuroepithelium, C4 to differentiating coelomic epithelia and C8 to proliferating cells. C1 and C3 represent most of the cells found in the coelomic epithelium of the anlage. In contrast, C0 represents a coelomic epithelial phenotype more closely associated with the mesentery than the intestine.

Trajectory Analysis: What populations are driving cell specification?

Among the many mysteries of the holothurian intestinal regeneration process is the identification of the precursor cells. In simple terms, what are the cells from which all nascent cells derive from? We performed a trajectory analysis of the data to provide some insights into this issue. This type of analysis is usually performed with samples at different stages or time points. However, we considered it feasible to conduct this analysis because in the 9-dpe regenerating anlage/mesentery we find cells at various stages of differentiation. These cells could provide crucial information on how the cell populations are associated with each other. To address this, we initially employed RNA velocity analysis. This method describes the temporal dynamics of gene expression based on the relative abundances of spliced and unspliced mRNA across cell populations.

Our initial velocity analysis on all the clusters and samples (Fig 9A-B), provided three main results. First, the direction of arrows in our UMAP shows them flowing towards C5, C9, and C4. These arrows do not point towards any other cluster; thus, they are terminal arrows. Second, while arrows from C8 are not terminal, they are directed towards C1. This suggests that cells in C1 provide cells for the growth of the anlage via proliferation. Thus, these results support C5, C9 and C4 as terminal cell clusters which we have described as muscle, neuroepithelial, and the nascent coelomic epithelium cells, respectively. Third, velocity embedding shows shorter arrows that point from clusters 0 and 1 towards terminal populations previously described. Therefore, cells of C0 and C1 are not undergoing significant transcriptional changes. The RNA velocity results of the rest of the clusters are less interesting as they do not show directions towards any other clusters, mainly because of their individuality within the UMAP. However, it is interesting that arrows of the mesenchymal cell populations show distinct directions and lengths. Based on the results, it seems that portions of both mesenchymal clusters (C2 and C7) have gone or are undergoing more extensive differentiation changes.

Trajectory analysis of cell populations from the regenerating intestinal tissue.

(A) UMAP plot of all identified clusters. (B) RNA velocity embedded in UMAP of all main clusters. (C) UMAP of re-clustering of cells from C0, C1, C3, C4, C5, and C9. (D) RNA velocity analysis results from the subset from panel (C). (E) Jitter plot of Slingshot pseudotime of cells from C0, C1, C3, C4, C5 and C9. Pseudotime resulted in two lineages one containing C5 and the other C 9. (F) UMAP of re-clustered anlage cells corresponding to C0, C1, C3, and C4(G) RNA velocity results from the cluster results from panel F. (H) UMAP of panel F plot overlayed with pseudotime results of Slingshot. Color represent pseudotime values from 0 (blue) to 60 (yellow). (I) Jitter plot showing the Slingshot pseudotiume of cells from each cell cluster.

We then re-clustered C0, C1, and C3 cells, along with the differentiating cell populations (C4, C5, and C9) to better understand their relationship (Fig 9C). The velocity assessment of these newly clustered populations resulted in a similar pattern (Fig 9D). Here, we can see that the differentiating cells (C4, C5, and C9) have long arrows suggesting that these cells are going through an advanced stage of transcriptional change compared to others to some extent, C4. Moreover, it seems that C0 has a closer relationship to cells of C5 and C9 and that some cells of these clusters could potentially differentiate into cells of C4 (the coelomic epithelium cells). In this case, to complement and confirm our RNA velocity interpretation, we also performed a pseudotime analysis using Slingshot, which relies on the expression data of each cluster. This analysis showed C1 in an earlier pseudotime than C3, C4 and C0 in the resulting two lineages (Fig 9D). The resulting lineages differed by the terminal clusters, one containing C5 (muscle) and the other C9 (neuroepithelial). Thus, it supports what we have already visualized on the RNA velocity embeddings.

The results described so far show that C0, C1 and C4 are cells in distinct differentiation states, but we wanted to have a clearer view of the cell clusters that potentially have an essential role in the regeneration process. For this, we made another subset of cells that corresponded to C0, C1, C3 and C4, but uniquely from cells of the anlage (Fig 9F). The rationale was that cells from the mesentery are certainly at a different state from those of the anlage and thus could interfere with the pseudotime of cells from the anlage. The RNA velocity analysis using this subset strengthened our previous inferences. First, C1 seems to be the least dedifferentiated cell cluster, whose future state will be cells of C3 and part of the population of C4. Second, that C0 seems to be in a specialized state of differentiation that has a relationship to C4 (Fig 9G). This would explain the relationship of this cluster to that of the differentiating cells of C5 and C9. Interestingly, portions of C1, C0 and C4 appear to be in an advance process of differentiation based on their longer arrows compared to C3 and another portion of C4 close to C3 (Fig 9G). The Slingshot analysis of the anlage cells from C0, C1, C3 and C4 revealed that the pseudotime starts at a point of convergence that contains a portion of cells from C1 and C0. Yet, it further supports the cells from C1 as the least differentiated (Fig 9H). C1 is then followed by cells of C3, C4 and lastly C0, which for the most part seems to be at a more advanced differentiation state with a closer relationship to differentiating cells (Fig 9I).

Discussion

In this study we employed scRNA seq and HCR-FISH techniques to examine the cellular phenotypes in regenerating intestinal tissues of the sea cucumber H. glaberrima. These techniques have been seldom used to characterize echinoderm cells, and the few studies available are mainly limited to embryonic stages of sea urchins (Strongylocentrotus purpuratus, Lytechinus variegatus) and the sea stars (Patiria miniata) (Cocurullo et al., 2023; Foster et al., 2022; Meyer et al., 2023; Paganos et al., 2022a, 2022b, 2021; Tominaga et al., 2023). Nonetheless, these studies provide an excellent description of the cell population and dynamics arising from major germ lines during echinoderm development. We now apply the same techniques to explore cell phenotypes involved in intestinal regeneration in holothurians. This is, to our knowledge, the first-time scRNA-seq and HCR-FISH have been used in adult echinoderms to analyze the cellular and molecular basis of their amazing regenerative properties. This research integrates the extensive cellular and molecular information on intestinal regeneration in holothurians collected over the past two decades, offering a comprehensive view of the cellular phenotypes and molecular changes involved.

Cell types and differentiation stages in the 9-dpe regenerating intestine

Our study has focused on the description of cells present in the 9-dpe regenerating intestine of the sea cucumber, where cells are known to have gone through dedifferentiation and are, at this time-point, in the process of differentiating into specialized cells (García-Arrarás et al., 2011). Based on our analysis we have identified 13 distinct populations that form part of the regenerating intestinal mesentery and anlage (Fig 1B). Among these clusters we have described the clusters corresponding to the coelomocyte, coelomic epithelium, and mesenchyme cell types (Fig 3). Thus, we can now use the information that our laboratory and others have gathered to pinpoint the cellular mediators and their activity during regeneration. The coelomic epithelia, for example, is highly influential in the regeneration process as it is the site where major cellular events occur, particularly cell division, dedifferentiation, and differentiation (García-Arrarás et al., 2011). Moreover, the mesenchyme ECM is known to undergo remodeling during the regeneration process, which is also critical for the proper growth of the new tissue (Quiñones et al., 2002). However, while much of the events have been described using microscopic and histological tools, we need a more comprehensive understanding of the transcriptomic characteristics of specific cell populations involved in the regeneration process.

The sea cucumber contains various mesenchymal and coelomocyte populations

Amongst the phenotypes identified are clusters of mesenchymal (C2 and C7) and coelomocyte cell populations (C6, C10, C11, and C12). The mesenchymal populations demonstrated a unique expression of ERG and ETS-1, markers of mesenchymal populations in other echinoderms. ETS-1 has long been associated with developmental processes of mesenchymal formation in sea urchin (Koga et al., 2010; Rizzo et al., 2006). In other studies, ERG, a member of the ETS gene families, has been found to be necessary for controlling mesenchymal identity and differentiation (Cox et al., 2014; Mochmann et al., 2013). Further analyses of their individual marker genes suggest that these cells might be involved in EMT. For instance, C2 expresses TIMP3, a metalloproteinase inhibitor that aids in the extracellular matrix remodeling (Dewing et al., 2020), and NPNT, a gene reported to be related to development and cancer processes (Magnussen et al., 2021). These genes are also crucial for cells undergoing EMT as the cells need to detach from the other cells and the basal lamina that forms the epithelium. Comparatively, the expression profile of C7 with genes such as PA21B (phospholipase A2), TMPS9 (transmembrane protease serine 9), SEPP1, DMBT1, ITIH3, and SVPE1 and its GO results suggest this mesenchymal population is undergoing different processes. Based on these contrasting expression profiles, we propose that C2 corresponds to cells that have recently undergone EMT from the coelomic epithelium and eventually differentiate into a more specialized phenotype (C7). Our pseudotime results further support this developmental transition as cells in C7 appears to be in a more advanced stage when compared to those of C2 (S6 Fig). Therefore, these two populations correspond to the first transcriptomic description of mesenchymal phenotypes reported in sea cucumber regenerating intestine.

Our dataset contains four distinct populations that we have characterized as coelomocytes. The coelomocyte populations reported in different holothuroid species ranged between 4 to 6 distinct types (Hetzel, 1963; Ramírez-Gómez et al., 2010; Xing et al., 2008). Studies from our laboratory previously revealed four different coelomocyte populations in H. glaberrima, distinguished by their morphology, histochemistry, and phagocytic activity. These were lymphocytes, phagocytes, spherulocytes and a population named “giant cells” (Ramírez-Gómez et al., 2010). The distinctive gene expression profile of each of the coelomocytes that we have identified can provide insights into the differences in their role as immune/circulating cells within the sea cucumber. For example, the C6 marker gene fibrinogen-like protein, which is part of a protein family known as FREP, makes this population of great interest.

Mainly because these molecules have been vastly studied across invertebrates, and multiple immune roles have been proposed, including phagocyte recognition and encapsulation (Hanington and Zhang, 2011). A distinct example is that of C10, where the expression of HCK, FER and ITF8 markers suggest this might be a macrophage-like activity (Chen et al., 2023; Dolgachev et al., 2018; Shuttleworth, 2018). Specifically, HCK, a member of Src family kinases (SFK), has been closely related to macrophage activation and polarization (Bhattacharjee et al., 2011; Poh et al., 2015). Interestingly, recent studies in A. japonicus found that an Src homolog mediates the phagocytosis of Vibrio splendidus, which further supports C10 immune identity (Wan et al., 2022). Thus, to our knowledge, this would be the first report of the expression profile of distinct coelomocyte populations in an adult echinoderm species, setting up the stage for integrating these populations with the previously described ones.

The coelomic epithelia of the intestinal anlage is composed of a heterogenous population of cells

Our research findings align with previous microscopic descriptions of cells in the normal and regenerating mesothelium and coelomic epithelia. In our data, we most easily identify the cell population that forms the cluster exhibiting a proliferative phenotype (C8). Proliferative cells in the regenerating intestine are primarily localized within the coelomic epithelium of the anlage (García-Arrarás et al., 2011, 1998; Reyes-Rivera et al., 2024), and the gene expression profile documented here unequivocally identifies these as proliferative.

A second population of cells that can be well correlated to previously described cells is that with a muscle cell phenotype (C5). The evidence suggests that this cell population represents those cells from the coelomic epithelium that are differentiating into myocytes. This evidence includes: (1) the top-expressed genes by the cells in this cluster are all muscle-associated genes; (2) the top-enriched terms are all related to muscle morphogenesis; (3) the cluster is mostly composed of cells that come from the anlage where muscle formation is known to be taking place at this stage (Murray and García-Arrarás, 2004); (4) differentiated muscle cells were not dissociated by the enzymatic procedure strongly suggesting that the muscle cells that we have identified in our data are those that are in a differentiation process, rather than fully differentiated cells closer to the body wall; (5) the cells in this cluster were identified by HCR-FISH of MYH7. These cells are localized toward the basal region of the coelomic epithelium, the region where the differentiating myocytes are known to be present (Murray and García-Arrarás, 2004). Moreover, among the top expressed genes of this cluster is Troponin I (TNNI1), which has also been reported to be highly expressed in muscle precursor cells of the sea star embryo and immature cardiomyocytes of the chicken (Mantri et al., 2021; Tominaga et al., 2023). The high differentiation activity of these cells is also supported by our pseudotime analysis, where distinct cells within the cluster are in individual differentiation states (Fig 9D&E). Further analysis of these populations could allow us to understand the transcriptional changes these cells undergo to become fully specialized muscle cells.

An additional population in our dataset, is the neuroepithelial population (C9). This population has STARD10, among its top expressed genes, known to be a phospholipid transfer protein present in a neural cell population localized in the coelomic epithelium of the intestine (García-Arrarás et al., 2019; Rosado-Olivieri et al., 2017). Furthermore, cells of this cluster have high expression of other genes reported to be expressed by neuronal cells of regenerative and developmental tissues, such as beta-tubulin in developing human gut and sea urchin larva, and synapsin in planaria regenerating tissue (Cocurullo et al., 2023; Elmentaite et al., 2020; King et al., 2024). Interestingly, an earlier study showing that beta-tubulin positive cells arise from dedifferentiated cells of the regenerating intestinal tissue of the sea cucumber (García-Arrarás et al., 2019) also suggests that similar to the muscle population, these cells are neuroepithelial cells that are differentiating rather than fully specialized. Further support for this theory lies in the main contribution to C9 coming from anlage cells and their terminal differentiation state observed in the pseudotime analysis (Fig 1D & 10D).

The remaining four clusters of the coelomic epithelium supra-cluster are more challenging to characterize. Nonetheless we will explore some hypotheses regarding their cellular phenotypes. The identity of C4 was put forward based on its close transcriptional and pseudotime correlation with populations of the coelomic epithelia (Figs 3 and 9). Additionally, in our in situ hybridization, its marker gene SC6A5, is mainly expressed by the coelomic epithelial cells of the normal intestine, strongly suggesting that the cells expressing this gene in the anlage are those that will become the coelomic epithelial cells of the regenerated organ (Fig 7).

C0 remains an intriguing cell population. On one hand it appears to be closely related, by its gene expression, to other coelomic epithelial populations (Fig 3). Thus, it may represent an unknown cell population or cell stage. The in situ expression suggest that this cluster represents the cellular population of the mesentery or peritoneal coelomic epithelium. In this case we might be evidencing differences in the coelomic epithelium of the mesentery (exemplified by cluster 0) from those of the coelomic epithelium of the intestine (exemplified by cluster 4). Future experiments will be needed to address this controversy.

Lastly, C1 and C3 share many common genes, and represent distinct populations but are still closely associated, regardless of the clustering parameters or the statistical assessment performed. Although we considered combining them into a single cluster, we ultimately decided against it, as they likely represent different stages of cell development or plasticity in the regenerating intestine. One-to-one comparisons revealed that C1 expressed various ribosomal gene markers, while C3 expressed specific genes such as EPHA4 (ephrin type-A receptor 4) and LRIG3, indicating distinct transcriptomic states. Additionally, all trajectory analyses revealed that C1 cells appear to give rise to C3 cells.

Further characterization of the least differentiated cells, C1 and C3, within the anlage coelomic epithelium suggest that these cells probably serve as cellular precursors to differentiating cells. The evidence from pseudotime analyses and gene expression of C1 and C3 strongly support this conclusion. These cells appear to exhibit pluripotency with the potential to form muscle, neurons, coelomic epithelia and mesenchymal cells in the regenerating intestine. Their gene expression profiles include markers from gene families associated with embryonic development such as Hox, zinc fingers, basic helix-loop-helix and others, some of which are linked to stemness and pluripotency. For example, HSP90A, a molecular chaperone essential for stem cell pluripotency, and markers like HES1 and TGFb, essential for maintaining stem cell proliferation, are present in these cell populations (Aztekin, 2021; Mishra et al., 2005). Moreover, the expression of PIWL1 (piwi-like protein 1) and YAP1 further support the classification of these clusters as precursor cells of the intestinal anlage. In hydra, piwi-like molecules are exclusively localized in stem/progenitor cells (Juliano et al., 2014), and in mice, YAP1 is expressed only in multipotent cells during intestinal epithelium regeneration, vital for their emergence (Ayyaz et al., 2019). Ultimately, trajectory analysis also supports C1 as a precursor cell population, as the dividing cells (C8) appear to give rise to C1, which then progresses toward C3. This analysis aligns with evidence of a proliferation center in the coelomic epithelia that provides precursor cells essential for the growth of the anlage.

C1 and C3 cells are probably dedifferentiated cells from the mesentery mesothelium, retaining markers common to all epithelial cells. Similar to what is known in other highly regenerative organisms, these dedifferentiated cells can proliferate and then re-differentiate into the cells of the new organ. Future experiments will determine whether they retain the memory of their previous phenotype (muscle or coelomic epithelium) or are completely pluripotent.

The coelomic epithelium of the intestinal anlage is pluripotent

Regardless of which cell population is responsible for giving rise to the cells of the regenerating intestine, our study reveals that the coelomic epithelium, as a tissue layer, is pluripotent. Thus, it is capable of giving rise to various cell types. Microscopy studies across different echinoderm species have consistently suggested that the coelomic epithelium can differentiate into various cell types. Among these are the formation of muscle cells and neurons in sea cucumbers (Yu Dolmatov et al., 1996), mesenchymal cells in brittle star (Piovani et al., 2021), and even immune cells (coelomocytes) in sea star (Byrne et al., 2020; Sharlaimova et al., 2021). In some species, the coelomic epithelium has even been proposed to transdifferentiate into intestinal luminal cells (Mashanov et al., 2005).

The direct association of pluripotency with the coelomic epithelium is evident in both the homeostatic regenerative processes that maintain cellular numbers and in the regenerative responses to injury or autotomy. In some cases, pluripotency is directly associated with the coelomic epithelium of the regeneration anlage. This is exemplified in the brittle star Marthasterias glacialis, where it has been suggested that the coelomic epithelium is not only involved in arm regeneration but also serves as a source of immune cells (Guatelli et al., 2022). This report indicated that “residential stem cells” in the regenerating arm of the brittle star originate from the coelomic epithelia (Candia-Carnevali et al., 2009). Similarly, in the sea cucumber Holothuria forskali, the injured mesothelial layer has been identified as the source of undifferentiated cells that differentiate into the cells of the growing organ and into phagocytic cells, now recognized as coelomocytes (Vandenspiegel et al., 2000).

Our scRNA-seq provides convincing evidence that various cell populations originate from the coelomic epithelia of the sea cucumber anlage, offering a more detailed view of the differentiation process. These cell populations, now classified by their gene expression data, correlate well with those previously characterized by microscopy descriptions. We have strong evidence for muscular, proliferating and neuroepithelial cells (C5, C8, and C9) arising from the coelomic epithelia of the anlage. We also show evidence of a differentiating cell population corresponding to the mature intestine coelomic epithelium (C4). Furthermore, our data suggest the presence of a mesenchymal population arising from the coelomic epithelium of the anlage through EMT (García-Arrarás et al., 2011).

Of particular interest, our study reveals that the mesentery coelomic epithelia population (C0) exhibits localized expression of SAA1, an immune response related gene. This finding, in line with previous reports of SAA1 localization in the coelomic epithelium of the regenerating intestinal tissue (Santiago et al., 2000) (S5 Fig), raises the intriguing possibility that the coelomic epithelium of the sea cucumber regenerating intestine could also be the source of coelomocytes, as suggested by Guatelli (2022) (Guatelli et al., 2022). This hypothesis is further supported by the observation that in the sea star Asteria rubens, coelomocytes arise from its coelomic epithelium (Vanden Bossche and Jangoux, 19976). However, our current results, while suggestive, do not yet provide conclusive evidence to support this in the sea cucumber, at least not in the coelomic epithelia of the regenerating intestine or the 9-dpe stage.

Currently, there is no definitive evidence identifying the precursor cells that give rise to these cell populations in echinoderms. However, given our gene expression data and the interactions observed among the cell populations, we postulate that cells from C1 stand as the precursor cell population from which the rest of the cells in the coelomic epithelium arise. Granted, much more experimental evidence will be needed to arrive at this conclusion.

Nonetheless, it provides a focus on the most likely, and probably most intriguing cell population. It also provides the path to explore multiple questions that arise from our results. Can C1 be subdivided into other cell populations? Are the cells in this cluster pluripotent? Do these cells originate from the same cells via dedifferentiation? These and many other questions will indeed be tested in future experiments.

Our data has allowed us to construct a model of the cell populations we identified in the 9-dpe intestinal anlage (Fig 10). This model presents the coelomic epithelium of the anlage as a heterogeneous layer of cells comprising of six cell populations corresponding to distinct differentiation states. Among these are cell populations in the process of differentiation: muscle, neuroepithelium, and coelomic epithelium cells. We also propose the presence of undifferentiated and proliferating cell populations in the coelomic epithelia, which give rise to the cells in this layer. Underneath the coelomic epithelia are the mesenchymal cells, some of which may have originated from the coelomic epithelial layer via EMT. Not shown in our model is a population localized in the coelomic epithelium of the mesentery that we infer also plays an important role in the regeneration process of the intestine.

Model of cellular organization within the anlage of the regenerating intestine of the sea cucumber H. glaberrima at 9 dpe.

Schematic highlights the heterogeneity of the coelomic epithelia of the intestinal anlage and identified mesenchymal populations. The colors of cells and C# correlate with those used on the UMAP of Fig 1B.

The intestinal anlage as a regeneration blastema

The intestinal anlage is a particular regenerative structure that has long been described as a blastema-like structure due to its remarkable morphological resemblance to the “classical blastema” (García-Arrarás et al., 1998). As stated previously a blastema is usually described as a transient structure composed of a mass of proliferating undifferentiated cells. This structure is found at the site of injury and will give rise to the regenerated organ (Seifert and Muneoka, 2018). The blastema cells can originate from different lineages across species. For example, in amphibians, the blastemal cells originate from dedifferentiated muscle, cartilage, fibroblast and other tissues, while in the flatworm they originate from undifferentiated stem cells known as neoblasts (Baguñà, 2012; Globus et al., 1980; Wagner et al., 2011). The blastema is overlayed by a wound epidermis that is formed by a re-epithelization process following injury, which develops into a specialized wound epidermis that in amphibians is known as the apical epithelial cap (AEC) (Aztekin, 2021). Since its first description, more than 100 years ago, the blastema has been regarded as the best indicator of regeneration, and its presence is associated with tissues or organs that are highly regenerative. In fact, it has been proposed that the presence or absence of a blastema defines the regenerative success or failure of the regeneration process. However, as more regenerative species continue to be studied, the classical definition of a blastema has been reconsidered, moving towards its functional role rather than its histological or structural characteristics (Seifert and Muneoka, 2018).

The blastema of salamanders and newts, have long served as models to describe a blastema cellular and molecular properties (Globus et al., 1980; Scimone et al., 2022; Tajer et al., 2023). The sea cucumber intestinal anlage stands as a different structure in terms of the tissue compartmentalization of certain activities. Yet, when examined closely, much of the processes and signaling molecules described in the amphibian blastema also take place in the holothurian anlage, although their spatial occurrence might differ. For instance, like the blastema cells of the amphibians, the cells of the holothurian anlage coelomic epithelium are proliferative undifferentiated cells and originated via a dedifferentiation process. Moreover, some of the genes expressed by the amphibian blastema cells or the overlying AEC, such as Wnt, Tgf-beta, and Fgf, are also known to be expressed during sea cucumber regeneration (Auger et al., 2023; Zeng et al., 2023). In amphibians, some of these factors are thought to be released by cells of the AEC, serving as a way of modulating the blastema cell activity. For example, FGF1 is expressed in the cells of the AEC while FGF receptors were localized to blastema cells(Zenjari et al., 1996). Moreover, both FGF and Wnt signaling have been shown to be vital for the formation of the blastema (Makanae et al., 2014).

This expression of Fgf and Wnt and their possible functions correlates with what has been found in sea cucumbers. Not only are the same growth factors expressed in C1 and C3, but a recent study highlighted the role of FGF4 as a modulator of cell proliferation during the intestinal regeneration of the sea cucumber A. japonicus (Zeng et al., 2023). This group demonstrated that inhibiting FGF4 and its receptor, FGFR2, negatively affected cell proliferation of the mesothelial layer during intestinal regeneration. Similar results have been shown for Wnt, primarily a study by our laboratory demonstrating that Wnt pathway inhibition, either by pharmacological drugs or by RNAi caused a reduction in cell proliferation (Alicea-Delgado and García-Arrarás, 2021; Bello et al., 2020). Thus, in the sea cucumber coelomic epithelia of the anlage, cells are also involved in intercellular communications that modulate cellular dynamics through factors similar to those found in the AEC-blastema cell modulation.

The holothurian anlage and the amphibian blastema exhibit some differences. In the latter there is a clear separation of the AEC and the underlying blastema cells. In amphibians, the AEC has been shown to actively participate in the formation and maintenance of the blastemal cells, particularly in their proliferation and differentiation. In contrast, in holothurians, cells with both blastema and AEC characteristics can be found within the coelomic epithelium. In fact, both PRRX1 (paired mesoderm homeobox protein), a gene expressed by amphibian blastema cell precursors (Gerber et al., 2018; Lin et al., 2021), and HES1 (transcription factor HES-1), a marker gene of the amphibian AEC, are expressed by the holothurian coelomic epithelium (Aztekin, 2021) (Fig 3 and S5 Fig).

In conclusion, in all cases where it has been studied, the blastema appears as a dynamic structure where communication among its cells and the overlying epidermis or epithelium plays important roles in the regenerative process. In this context, the sea cucumber intestinal anlage could be considered a blastema since it uses similar mechanisms to fulfills the same role in the regeneration of the new organ. In this case, the cells of the coelomic epithelium would be those considered to be the blastemal cells.

Methods

Animals and Sample Collection

Two adult sea cucumbers were collected from the northeastern rocky region of Puerto Rico and were kept in aerated sea water for acclimatation prior to experiments. Evisceration of the sea cucumber was stimulated by intracoelomic injection of 0.35 M KCl as previously described (Reyes-Rivera et al., 2024). They were maintained in sea water aquaria for 9 days to undergo regeneration until dissection. Before the dissection, animals were anesthetized by immersion in ice-cold water for 45 minutes. Dissection was done through an initial dorsal incision which allowed exposition of the internal organs, upon which the growing rudiment (or anlage) and mesentery were dissected and separated from each other. Each tissue was kept in CMFSS on ice and treated separately during the tissue dissociation process.

Tissue Dissociation

Mesentery and rudiment tissues were digested for 15 minutes, in rocking shaker at room temperature with 1mL of 0.05% Trypsin/0.02% EDTA solution prepared in CMFSS (Bello et al., 2015). Digestions were quenched by adding 500µL of 0.2% BSA in CMFSS and then centrifuged for 2 minutes at 1500 rpm. The supernatants were discarded, and the pellets resuspended with 500µL of 0.04% BSA in CMFSS and the cells were gently separated by pipetting using glass pipets with fire blunted tips. The cell suspensions were filtered using a nylon cell strainer with 70mm mesh and aliquots were taken for cell counting. Cell viability was assessed through manual cell counting with hemocytometer to confirm that a viability higher than 90% was maintained. Sample suspensions were adjusted to 1000 cells/mL using CMFSS, as required for sequencing procedures.

Immuno- and cytochemistry

Fifty µL of dissociated cell sample were placed on polysine-treated slides and fixed with 50 µL of 4% paraformaldehyde and left to dry overnight. Slides were washed in PBS prior to use for immune and or cytochemistry. The methodology for the immunofluorescence techniques/preparation of slides was followed as published except for the time of PBS washes which were of 10 minutes instead of 15 minutes (Diaz-Balzac et al., 2007; Reyes-Rivera et al., 2024). The primary and secondary antibodies used are in S7 Table. Samples to be probed with fluorescent phalloidin were treated directly with Phalloidin-TRITC (1:1000; Sigma P1951) for 1h as described previously (Reyes-Rivera et al., 2024). DAPI was incorporated into the mounting medium as described previously (Reyes-Rivera et al., 2024). Cells were observed in a Nikon DS-Qi2 fluorescent microscope.

Single Cell RNA Sequencing and Data Analysis

Single cell libraries were prepared using the 10X Genomics Chromium Next GEM Single Cell 3’ Kit v3.1 and the Chromium 10X instrument, following the protocol from manufacturer (CG000204). Sequencing was carried out with Illumina NextSeq 2000 at the Sequencing and Genotyping Facility of the University of Puerto Rico Molecular Science Building.

The raw sequencing reads from all four samples (two mesentery and two anlage) were processed individually using Cell Ranger (v7.1.0) (Zheng et al., 2017) using the reference genome and gene models of H. glaberrima available at blastkit.hpcf.upr.edu/hglaberrima-v1. Gene models were annotated against human reference proteins from Uniprot and the entire Uniprot reference protein database. Gene IDs throughout the study correspond to the annotations with the human reference protein sequences from Unitprot.

Initial quality control was conducted in R (v4.3.2) using SoupX (v1.6.2) with default parameters to remove ambient RNA (Young and Behjati, 2020). Subsequent data processing was carried out using Seurat (v4.3.0.1) (Hao et al., 2024, 2021; Satija et al., 2015; Stuart et al., 2019). After data normalization and identifying variable features (n=2000), 22,970 anchors were identified using 20 dimensions. The datasets were integrated using the canonical correlation analysis (CCA) approach of Seurat, followed by data scaling for principal component analysis (PCA) (npcs = 50) and Uniform Manifold Approximation and Projection (UMAP) dimensional reduction technique (dims = 1:20).

Neighbors were then identified with 20 dimensions, leading to cluster identification at a resolution of 0.5. Various resolution parameters (0.8, 1.2, and 1.4) and dimensions (up to 35 in increments of 5) were tested before the final resolution value. The statistical analysis tool scSHC was employed to assess the probability of each cluster being unique (Grabski et al., 2023). Marker genes of each cluster or supra-clusters were identified using the FindMarkers() function in Seurat and mapped to the appropriate Uniprot IDs. Genes highlighted throughout the study where further validated using the NCBI non-redundant reference database and EchinoBase (Telmer et al., 2024) via BLAST.

Differential expression data across clusters was used to perform gene set enrichment analysis of Gene Ontology (gseGO) biological processes terms using clusterProfiler (v.10.0) (Yu et al., 2012) with a p-value cutoff of 0.05. For this analysis, BLASTp was used to map H. glaberrima gene models to human reference protein sequences from NCBI (GCF_000001405.40), facilitating the assignment of ENTREZ ID to the correct human GO terms in the AnnotationDbi (v1.64.1) human database (v2.1) (Pagès et al., 2024).

RNA velocity loom files were generated with velocyto (v0.17), which relied on genome-masked regions obtained from RepeatModeler (v2.0.5) and RepeatMasker (v4.1.5) (Smit et al., 2015; Smit and Hubley, 2015), along with the sea cucumber gene models and CellRanger dataset. These files were further analyzed with velocyto.R (v0.6) and SeuratWrappers (v0.2.0), using the ReadVelocity and RunVelocity functions. Pseudotime analysis for the distinct data subsets was conducted using Slingshot (v2.10.0) (Street et al., 2018). The code of the data analysis has been made available at GitHub (https://github.com/devneurolab/scRNAseq_Hglaberrima).

HCR-FISH

Regenerating intestines from animals eviscerated 8-9 days previously and intestines from non-eviscerated (controls) animals were collected and fixed in 4% (v/v) paraformaldehyde with phosphate-buffered saline (0.01M PBS; 0.138M NaCl; 0.0027M KCl; pH7.4) overnight at 4 ℃. Tissues were then, washed three times by PBS, and treated overnight with 40% saccharose prior to cutting in the cryostat. Sections (20 mm) were prepared in a cryostat (Leica CM1850), as previously published for immunohistochemistry (Reyes-Rivera et al., 2024). Gene spatial expression was determined by designing twelve (12) or twenty (24) set split probes for each gene marker. To decrease probe unspecific binding and increase its signal to background ratio, probes were designed as described by H. Choi and colleagues (Molecular Instruments) (Choi et al., 2018). To ensure probe gene target specificity, all nucleotide regions selected for probe design underwent extensive validation using the newly developed H. glaberrima genome and transcriptome alignment tool (https://blastkit.hpcf.upr.edu/hglaberrima-v1/) (Medina-Feliciano et al., 2021).

Probes were used at a final concentration of approximately 20-25 nM. Immediately after slide preparation, HCR-FISH v3 was carried out using a modified version of Molecular Instrument’s fresh fixed frozen tissues protocol. Probe hybridization was conducted overnight at 37 ℃, while DNA hairpin amplification (B1-546nm or B2-546nm) was done at 25℃ overnight with 3pmol of h1 and h2, respectively. Our protocol modification involved replacing Ethanol with Methanol during sample permeabilization. Also, the use of proteinase K was omitted, and Tween 20 at 0.1% was added to all PBS wash buffers. Once slides were prepared, image acquisition was attained using a Nikon DS-Qi2 fluorescent microscope. Positive control probes (PolyA) were used as signal calibrators to define background from positive signal during image analysis.

eLife Assessment

This study presents a dataset obtained through a single cell RNA-Sequencing of sea cucumber regenerating intestine 9 days post evisceration. The data were collected and analyzed using standard single cells analysis from n=2 adult sea cucumbers captured from the wild, which represents a useful resource for future studies. Although cell type validation is attempted, it is performed on samples from the same 2 animals (and not independent samples), rendering the validation incomplete. Further, the RNA localization images provided in the paper could benefit from improved spatial context, and many strong statements in the discussion should be better justified and supported by the presented data. With the validation part strengthened, this paper would be of interest to development and regeneration fields.

Reviewer #1 (Public Review):

Summary:

Medina-Feliciano et al. investigated the single cell transcriptomic profile of holoturian regenerating intestine following evisceration, a process used to expel their viscera in response to predation. Using single cell RNA-sequencing and standard analysis such as "Find cluster markers", "Enrichment analysis of Gene Ontology" and "RNA velocity", they identify 13 cell clusters and potential identity. Based merely on bioinformatic analysis they identified potentially proliferating clusters and potential trajectories of cell differentiation. This manuscript represents a useful dataset that can provide candidate cell types and cell markers for more in-depth functional analysis for gaining a better understanding of the holoturian intestine regeneration. The conclusions of this paper are supported only by bioinformatic analyses, since the in vivo validation through HCR does not sufficiently support them.

Strengths:

- The Authors are providing a single cell dataset obtained from sea cucumber regenerating their intestine. This represents a first fundamental step to an unbiased approach to better understand this regeneration process and the cellular dynamics taking part in it.

- The Authors run all the standard analyses providing the reader with a well digested set of information about cell clusters, potential cell types, potential functions and potential cell differentiation trajectories.

Weaknesses:

- The entire study is based on only 2 adult animals, that were used for both the single cell dataset and the HCR. Additionally, the animals were caught from the ocean preventing information about their age or their life history. This makes the n extremely small and reduces the confidence of the conclusions.

- All the fluorescent pictures present in this manuscript present red nuclei and green signals being not color-blind friendly. Additionally, many of the images lack sufficient quality to determine if the signal is real. Additional images of a control animal (not eviscerated) and of a negative control would help data interpretation. Finally, in many occasions a zoomed out image would help the reader to provide context and have a better understanding of where the signal is localized.

- The Authors frequently report the percentage of cells with a specific feature (either labelled or expressing a certain gene or belonging to a certain cluster). This number can be misleading since that is calculated after cell dissociation and additional procedures (such as staining or sequencing and dataset cleanup) that can heavily bias the ratio between cell types. Similarly, the Authors cannot compare cell percentage between anlage and mesentery samples since that can be affected by technical aspects related to cell dissociation, tissue composition and sequencing depth.

- The Authors decided to validate only a few clusters and in many cases there are no positive controls (such as specific localization, specific function, changes between control and regenerating animals, co-stain) that could actually validate the cluster identity and the specificity of the selected marker. There is no validation of the trajectory analysis and there is no validation of the proliferating cluster with H3P or BrdU stainings.

- It is not clear what is already known about holothurian intestine regeneration and what are the new findings in this manuscript. The Authors reference several papers throughout the whole result sectioning mentioning how the steps of regeneration, the proliferating cells, some of the markers and some of the cell composition of mesenteries and anlages was already known.

Reviewer #2 (Public Review):

Summary:

This research offers a comprehensive analysis of the regenerative process in sea cucumbers and builds upon decades of previous research. The approach involves a detailed examination using single-cell sequencing, making it a crucial reference paper while shedding new light on regeneration in this organism.

Strengths:

Detailed analysis of single-cell sequencing data and high-quality RNA localization images provide significant new insights into regeneration in sea cucumbers and, more broadly, in animals.

Weaknesses:

The spatial context of the RNA localization images is not well represented, making it difficult to understand how the schematic model was generated from the data. In addition, multiple strong statements in the conclusion should be better justified and connected to the data provided.

Reviewer #3 (Public Review):

Summary:

The authors have done a good job of creating a "resource" paper for the study of gut regeneration in sea cucumbers. They present a single-cell RNAseq atlas for the reconstitution of Holothuria glaberrima gut following self-evisceration in response to a potassium chloride injection. The authors provide data characterizing cellular populations and precursors of the regenerating anlage at 9 days post evisceration. As a "Tools and Resources" contribution to eLife, this work, with some revisions, could be appropriate. It will be impactful in the fields of regeneration, particularly in invertebrates, but also in comparative studies in other species, including evolutionary studies. Some of these comparative studies could extend to vertebrates and could therefore impact regenerative medicine in the future.

Strengths:

• Novel and useful information for a model organism and question for which this type of data has not yet been reported

• Single-cell gene expression data will be valuable for developing testable hypotheses in the future

• Marker genes for cell types provided to the field

• Interesting predictions about possible lineage relationships between cells during sea cucumber gut regeneration

Weaknesses:

• Possible theoretical advances regarding lineage trajectories of cells during sea cucumber gut regeneration, but the claims that can be made with this data alone are still predictive

• Better microscopy is needed for many figures to be convincing

• Some minor additions to the figures will help readers understand the data more clearly

Author response:

The following is the authors’ response to the original reviews.

Public Reviews:

Reviewer #1:

The entire study is based on only 2 adult animals, that were used for both the single cell dataset and the HCR. Additionally, the animals were caught from the ocean preventing information about their age or their life history. This makes the n extremely small and reduces the confidence of the conclusions.

This statement is incorrect. While the scRNAseq was indeed performed in two animals (n=2), the HCR-FISH was performed in 3-5 animals (depending on the probe used). These were different animals from those used for the scRNAseq. The number of animals used has now been included in the manuscript.

All the fluorescent pictures present in this manuscript present red nuclei and green signals being not color-blind friendly. Additionally, many of the images lack sufficient quality to determine if the signal is real. Additional images of a control animal (not eviscerated) and of a negative control would help data interpretation. Finally, in many occasions a zoomed out image would help the reader to provide context and have a better understanding of where the signal is localized.

Fluorescent photos have been changed to color-blind friendly colors. Diagrams, arrows and new photos have been included as to guide readers to the signal or labeling in cells. Controls for HCR-FISH and labeling in normal intestines have been included.

Reviewer #2:

The spatial context of the RNA localization images is not well represented, making it difficult to understand how the schematic model was generated from the data. In addition, multiple strong statements in the conclusion should be better justified and connected to the data provided.

As explained above we have made an effort to provide a better understanding of the cellular/tissue localization of the labeled cells. Similarly, we have revised the conclusions so that the statements made are well justified.

Reviewer #3:

Possible theoretical advances regarding lineage trajectories of cells during sea cucumber gut regeneration, but the claims that can be made with this data alone are still predictive.

We are conscious that the results from these lineage trajectories are still predictive and have emphasized this in the text. Nonetheless, they are important part of our analyses that provide the theoretical basis for future experiments.

Better microscopy is needed for many figures to be convincing. Some minor additions to the figures will help readers understand the data more clearly.

As explained above we have made an effort to provide a better understanding of the cellular/tissue localization of the labeled cells.

Recommendations for the authors:

Reviewer #1 (Recommendations For The Authors):

- Page 4, line 70-81: if the reader is not familiar with holothurian anatomy and regeneration process, this section can be complicated to fully understand. An illustration, together with clear definitions of mesothelium, coelomic epithelium, celothelium and luminal cells would help the reader.

A figure (now Figure 1) detailing the holothurian anatomy of normal and regenerating animals has been added. A figure detailing the intestinal regeneration process has also been included (S1).

- Page 5 line 92-104: this paragraph could be shortened. It would be more important to explain what the main question is the Authors would like to answer and why single cell would be the best technique to answer it, than listing previous studies that used scRNA-Seq.

The paragraph has been shortened and the focus has been shifted to the question of cellular components of regenerative tissues in holothurians.

- Page 6, line 125-127 and line 129-132: this belongs to the method section.

This information is now provided in the Materials and Methods section.

- Page 11, line 210-217: this belongs to the discussion.

This section has now been included in the Discussion.

- How many mesenteries are present in one animal?

This has now been included as part of Figure S1.

- In the methods there are no information about the quality of the dataset and the sequencing and the difference between the 2 samples coming from the 2 animals. How many cells from each sample and which is the coverage? The Authors provided this info only between mesentery and anlage but not between animals.

We have added additional information about the sequencing statistics in S4 Fig and S15 Table. Description has also been added in the methods in lines 922-926 under Single Cell RNA Sequencing and Data Analysis section.

- The result section "An in-depth analysis of the various cluster..." is particularly long and very repetitive. I would encourage to Authors to remove a lot of the details (list of genes and GO terms) that can be found in the figures and stressed only the most important elements that they will need to support their conclusions. Having full and abbreviated gene names and the long list of references makes the text difficult to read and it is challenging to identify the main point that the Authors are trying to highlight.

This section has been abbreviated.

- Figure 1: I would suggest adding a graph of holothurian anatomy before and after the evisceration to provide more context of the process we are looking at and remove 1C.

Information on the holothurian anatomy has been included in a new Fig 1 and in supplementary figure S1

- Figure 2: I would suggest removing this figure that is redundant with Figure 3 and several genes are not cluster specific. Figure 3 is doing a better job in showing similar concepts.

Figure 2 was removed and placed in the Supplement section.

- In figure 3 how were the 3 cell types defined? Was this done manually or through a bioinformatic analysis?

The cell definition was done following the analysis of the highly expressed transcripts and comparisons to what has been shown in the scientific literature.

- Figure 2O shows that one of the supra-cluster is made of C2, C7, C6 and C10. This contradicts the text page 9, line 195.

The transcript chosen for this figure gives the wrong idea that these 4 clusters are similar. We have now addressed this in the manuscript.

- Figure 4A and 4C: if these are representing a subset of Figure 3, they should be removed in one or the other. The same comment is valid also for Figures 5, 6 and 7. In general the manuscript is very redundant both in terms of Figures and text.

These are indeed subsets of Fig 3 that were added with the purpose of clarifying the findings, however, in view of the reviewer’s comment we have deleted the redundant information from all figures.

- Figure 9: since the panels are not in order, it is difficult to follow the flow of the figure. - All UMAP should have the number of the cluster on the UMAP itself instead of counting only on the color code in order to be color-blind friendly.

The figure has been modified and clusters are now identified in the UMAP by their number.

- Figure S1F seems acquired in very different conditions compared to the other images in the same figure.

Fig S1F (now S2 Fig) is an overlay of fluorescent immune-histochemistry (UV light detected) with “classical” toluidine blue labeling (visible light detected). This has now been explained in the figure legend.

- Table S7 is lacking some product numbers.

The toluidine blue product number has now been added to the table. The antibodies that lack product number correspond to antibodies generated in our lab and described in the references provided.

- The discussion is pretty long and partially redundant with the result section. I would encourage the Authors to shorten the text and shorten paragraphs that have repeating information. - It might be out of the scope of the Authors but the readers would benefit from having a manuscript that focuses more on the novel aspects discovered with the single-cell RNA-Seq and then have a review that will bring together all the literature published on this topic and integrating the single-cell data with everything that is known so far.

We have tried to shorten the discussion by eliminating redundant text.

Reviewer #2 (Recommendations For The Authors):

- An intriguing finding is the lack of significant difference in the cell clusters between the anlage and mesentery during regeneration. This discovery raises important questions about the regenerative process. The authors should provide a more detailed explanation of the implications of this finding. For example, does it suggest that both organs contribute equally to the regenerated tissues?

The lack of significant differences in the cell clusters between the anlage and the mesentery is somewhat surprising but can be explained by two different facts. First, we have previously shown that many of the cellular processes that take place in the anlage, including cell proliferation, apoptosis, dedifferentiation and ECM remodeling occur in a gradient that begins at the tip of the mesentery where the anlage forms and extends to various degrees into the mesentery. Similarly, migrating cells move along the connective tissue of the mesentery to the anlage. Thus, there is no clear partition of the two regions that would account for distinct cell populations associated with the regenerative stage. Second, the two cell populations that would have been found in the mesentery but not in the regenerating anlage, mature muscle and neurons, were not dissociated by our experimental protocol as to allow for their sequencing. Our current experiments are being done using single nuclei RNA sequencing to overcome this hurdle. This has now been included in the discussion.

- Proliferating cells are obviously important to the study of regeneration as it is assumed these form the regenerating tissue. The authors describe cluster 8 as the proliferative cells. Is there evidence of proliferation in other cell types or are these truly the only dividing cells? Is c8 of multiple cell types but the clustering algorithm picks up on the markers of cell division i.e. what happens if you mask cell division markers - does this cluster collapse into other cluster types? This is important as if there is only one truly proliferating cell type then this may be the origin of the regenerative tissues and is important for this study to know this.

As the reviewer highlights, we also believe this to be an important aspect to discuss. We have addressed this in the manuscript discussion with the following: “Our data suggest that there appears to be a specific population of only proliferative cells (C8) characterized by a large number of cell proliferation genes, which can be visualized by the top genes shown in Fig 3. These cell proliferation genes are specific to C8, with minimum representation in other populations. Interestingly, as mentioned before C8 expresses at lower levels many of the genes of other coelomic epithelium populations. Nevertheless, even if we mask the top 38 proliferation genes (not shown), this cluster is maintained as an independent cluster, suggesting that its identity is conferred by a complex transcriptomic profile rather than only a few proliferation-related genes. Therefore, the identity and potential role of C8 could be further described by two distinct alternatives: (1) cells of C8 could be an intermediate state between the anlage precursor cells (discussed below) and the specialized cell populations or (2) cells of C8 are the source of the anlage precursor populations from which all other populations arise. The pseudotime data is certainly complex and challenging to interpret with our current dataset, yet the RNA velocity analysis showed in Fig 11B would suggests that cells of C8 transition into the anlage precursor populations, rather than being an intermediate state. This is also supported by the Slingshot pseudotime analysis that incorporates C8 (S13 Fig).

Nevertheless, additional experiments are needed to confirm this hypothesis.”

- The schematic model presented in Fig 10 is essential for clarifying the paper's findings and will provide a crucial baseline model for future research. However, the comparison of the data shown in the HCR figures with the schematic is challenging due to the lack of spatial context in the HCR figures. The authors should find a way to provide better context in the figures, such as providing two-color in situ images to compare spatial relationships of cell types and/or including lower resolution and side-by-side fluorescent and bright field images if possible.

The figure has been modified to explain the spatial arrangement of the tissues.

The authors make several strong statements in the discussion that weren't well connected to the findings in the data. Specifically:

“Regardless of which cell population is responsible for giving rise to the cells of the regenerating intestine, our study reveals that the coelomic epithelium, as a tissue layer, is pluripotent.”

This has now been expanded to better explain the statement.

738 “…we postulate that cells from C1 stand as the precursor cell population from which the rest of the cells in the coelomic epithelium arise”.

This has now been expanded to better explain the statement.

748 “differentiation: muscle, neuroepithelium, and coelomic epithelium cells. We also propose the presence of undifferentiated and proliferating cell populations in the coelomic epithelia, which give rise to the cells in this layer…”

This has now been expanded to better explain the statement.

777 “amphibians, the cells of the holothurian anlage coelomic epithelium are proliferative undifferentiated cells and originated via a dedifferentiation process…”

This has now been expanded to better explain the statement.

Reviewer #3 (Recommendations For The Authors):

Specific questions:

- Is there any way to systematically compare these cells to evolutionarily-diverged cells in distant relatives to sea cucumbers? Or even on a case-by-case basis? For example, is there evidence for any of these transitory cell types to have correlate(s) in vertebrate gut regeneration?

This is a most interesting question but one that is perhaps a bit premature to answer due to multiple reasons. First, most of the studies in vertebrates focus on the regeneration of the luminal epithelium, a layer that we are not studying in our system since it appears later in the regeneration process. Second, there is still too little data from adult echinoderms to fully comprehend which cells are cell orthologues to vertebrates. Third, we are only analyzing one regenerative stage. It is our hope that this is just the start of a full description of what cell types/stages are found and how they function in regeneration and that this will lead us to identify the cellular orthologues among animal species.

Major revisions:

- If lineage tracing is within the scope of this paper, it would provide more definitive evidence to the conclusions made about the precursor populations of the regenerating anlage.

Response: This is certainly one of the next steps, however at present, it is not possible due to technical limitations.

Minor revisions:

- Line 47: "for decades" even longer! Could the authors also cite some other amphibians, such as other salamanders (newts) and larval frogs?

References have been added.

- Line 85: "specially"-could authors potentially change to "specifically"

Corrected

- Line 122: Authors should add the full words of what these abbreviations stand for in the caption for Figure 1 or in Figure 1A itself.

Corrected

- Lines 153: What conclusions are the authors trying to make from one type of tubulin presence compared to the others? It's unclear from the text.

The authors are not trying to reach any particular conclusion. They are just stating what was found using several markers, and the possibility that what might be viewed first hand as a single cell population might be more heterogenous. Although the tubulin-type information might not be relevant for the conclusions in the present manuscript, it might be important for future work on the cell types involved in the regeneration process.

- Line 226: Could the authors clarify if "WNT9" is "WNT9a". Figure 3 lists WNT9a but authors refer to WNT9 in the text.

The gene names in Fig 3 are based on the human identifiers. H. glaberrima only has one sequence of Wnt9 (Auger et al. 2023) and this sequence shares the highest similarity to human Wnt9a, thus the name in the list. We have now identified the gene as Wnt9 to avoid confusion.

- Lines 236-237: Can authors rule out that some immune cells might infiltrate the mesenchymal population?

No, this cannot be ruled out. In fact, we believe that most of the immune cells found in our scRNA-seq are indeed cells that have infiltrated the anlage and are part of the mesenchyma. This has been reported by us previously (see Garcia-Arraras et al. 2006). We have now included this in the text.

- Line 452-453: The over-representation of ribosomal genes not shown. Would it be possible to show this information in the supplementary figures?

The sentence has been modified, the data is being prepared as part of a separate publication that focuses on the ribosomal genes.

- Line 480: Could authors clarify if it's WNT9a or just WNT9?

It is indeed Wnt9. See previous response above.

- Line 500: In future experiments, it would be interesting to compare to populations at different timepoints in order see how the populations are changing or if certain precursors are activated at different times.

We fully agree with the reviewer. These are ongoing experiments or are part of new grant proposals.

- Line 567-568: Choosing 9-dpe allowed for 13 clusters, but do authors expect a different number of clusters at different timepoints as things become more terminally differentiated?

Definitely, we believe that clusters related to the different regenerative stages of cells can be found by looking at earlier or later regeneration stages of the organ. A clear example is that if the experiment is done at 14-dpe, when the lumen is forming, cells related to luminal epithelium populations will appear. It is also possible that different immune cells will be associated with the different regeneration stages.

- Line 653: References Figure 10D (not in this manuscript). Are authors referring to only 1D or 9D or an old draft figure number?

As the reviewer correctly points out, this was a mistake where the reference is to a previous draft. It has now been corrected.

- Line 701: "our study reveals that the coelomic epithelium, as a tissue layer, is pluripotent." Phrasing may be better as referring to the cell population making up the tissue layer as pluripotent/multipotent or that the cells it contains would likely be pluripotent or multipotent. Additionally, lineage tracing may be needed to definitively demonstrate this.

This has been modified.

- Line 808: The authors may make a more accurate conclusion by saying that the characteristics are similar to blastemas or behave like a blastema rather than it is blastema. There is ambiguity about the meaning of this term in the field, but most researchers seem to currently have in mind that the "blastema" definition includes a discrete spatial organization of cells, and here these cells are much more spread out. This could be a good opportunity for the authors to engage in this dialogue, perhaps parsing out the nuances of what a "blastema" is, what the term has traditionally referred to, and how we might consider updating this term or at least re-framing the terminology to be inclusive of functions that "blastemas" have traditionally had in the literature and how they may be dispersed over geographical space in an organism more so than the more rigid, geographically-restricted definition many researchers have in mind. However, if the authors choose to elaborate on these issues, those elaborations do belong in the discussion, and the more provisional terminology we mention here could be used throughout the paper until that element of the revised discussion is presented. We would welcome the authors to do this as a way to point the field in this direction as this is also how we view the matter. For example, some of the genes whose expression has been observed to be enriched following removal of brain tissue in axolotls (such as kazald2, Lust et al.), are also upregulated in traditional blastemas, for instance, in the limb, but we appreciate that the expression domain may not be as localized as in a limb blastema. Additionally, since there is now evidence that some aspects of progenitor cell activation even in limb regeneration extend far beyond the local site of amputation injury (Johnson et al., Payzin-Dogru et al.), there is an opportunity to connect the dots and make the claim that there could be more dispersion of "blastema function" than previously appreciated in the field. Diving a bit more into these nuances may also enable better conceptual framework of how blastema function may evolve across vast evolutionary time and between different injury contexts in super-regenerative organisms.

We have followed the reviewer’s suggestion and stated that the holothurian anlage behaves as a blastema. Though we would love to elaborate on the blastema topic, as suggested by the reviewer, we believe that it would extend the discussion too much and that the topic might be better served in a different publication.

- In the discussion, it would be important not to leave the reader with the impression that all amphibian blastema cells originate via dedifferentiation. This is not the case. For example, in axolotls (Sandoval-Guzman et al.) and in larval/juvenile newts, muscle progenitors within the blastema structure have been shown to originate from muscle satellite cells, a kind of stem cell, in stump tissues (while adult newts use dedifferentiation of myofibers to generate muscle progenitors in the blastema). Most cell lineages simply have not been evaluated in the level of detail that would be required to definitively conclude one way or the other, and the door is open for a more substantial contribution from stem cell populations than previously appreciated especially because new tools exist to detect and study them. Providing the reader with a more nuanced view of this situation will not negatively impact the findings in this paper, but it will show that there is biological complexity still waiting to be discovered and that we don't have all the answers at this point.

This has now been corrected.

Figures: Overall, the figures need minor work.

- Figure 1A: Can the authors draw a smaller, full-body cartoon and feature the current high-mag cartoon as an inset to that? Can they label the axes and make it clear how the geometry works here?

Fig 1 has been re-done and now is split into Fig 1 and Fig 2.

- Figure 1B: Can the authors label the UMAP with cluster identities on the map itself? This will make it easier to identify each cluster (especially to make sure cluster 11 is easier to find).

This has been corrected.

- Figure 2: Could the authors put boxes/clearly distinguish panel labels around each cluster (AO), so that there are clear boundaries?

Fig 2 has been moved to Supplement, following another reviewer recommendation.

- "Gene identifiers starting with "g" correspond to uncharacterized gene models of H. glaberrima." - The sentence is from another figure caption but this figure would benefit from having this sentence in the figure caption as well.

This has been added to other figures as suggested.

- Figure 3A: Can the authors potentially bold, highlight, or underline genes you discuss in text, so it's easier for the reader to reference?

This has been added as suggested.

- Figure 3C: Can the authors please label the cell types directly on the UMAP here as well?

The changes were made following the reviewer’s recommendation.

- Figure 4D-E: There's not much context here to determine if this HCR-FISH validation can tell us anything about these cells besides some of them appear to be there. Do authors expect the coelomocyte morphology to look different in regenerating/injured tissue versus normal animals? Can the authors provide some double in situs, as well as some lower-magnification views showing where the higher-magnification insets are located? Is there any spatial pattern to where these cells are found? Counter stains would be helpful.

- Figure 6C: If clusters C5, C8, C9 are part of the coelomic epithelium, then authors could show a smaller diagram above with blue and grey to show types and then show clusters separately to help get their point across better.

- Figure 6G: This image appears to have high background- would it be possible for authors to repeat phalloidin stain or reimage with a lower exposure/gain. Additionally, imaging with Zstacks would help to obtain maximum intensity projections. It would greatly aid the reader if each image was labeled with HCR probes/antibodies that have been applied to the sample.

- Figure 7E: The cells appear to be out of focus and have high background. Additionally, they are lacking the speckled appearance expected to be seen with HCR-FISH. Would it be possible for authors to collect another image utilizing z-stacks?

HCR-FISH figures identifying the gene expression characteristic of cell clusters have been modified following the reviewer’s concerns. The changes include:

(1) Additional clusters have been verified with probes to gene identifiers. These include clusters 8, 9 and 12.

(2) Redundant information has been removed.

(3) Colors have been changed to make figures friendlier to color-impaired readers.

(4) Spatial context has been added or identified.

(5) In some cases, improved photos have been added

(6) Better labels have been included

(7) When necessary individual photos used for the overlay have been included.

- Figure 9A: Could authors add cluster labels onto UMAP directly?

This change was made to Fig 2A. UMAP in Fig 9A is the same and used just as reference of the subset.

- Figure 10: It could be useful if authors put a small map of the sea cucumber like in other images so that readers know where in the anlage this zoomed in model represents.

Added as suggested by the reviewer.

- Supplementary figure 1F: Could authors add an arrow to the dark cell that's being pointed out?

Changed made as suggested by the reviewer.

- Supplementary figure 1: Could authors label clearly what color is labeled with what marker?

Changed made as suggested by the reviewer.

Version 2

Abstract

Abstract

In holothurians, the regenerative process following evisceration involves the development of a “rudiment” or “anlage” at the injured end of the mesentery. This regenerating anlage plays a pivotal role in the formation of a new intestine. Despite its significance, our understanding of the molecular characteristics inherent to the constituent cells of this structure has remained limited. To address this gap, we employed state-of-the-art scRNA-seq and HCR-FISH analyses to discern the distinct cellular populations associated with the regeneration anlage. Through this approach, we successfully identified thirteen distinct cell clusters. Among these, two clusters exhibit characteristics consistent with putative mesenchymal cells, while another four show features akin to coelomocyte cell populations. The remaining seven cell clusters collectively form a large group encompassing the coelomic epithelium of the regenerating anlage and mesentery. Within this large group of clusters, we recognized previously documented cell populations such as muscle precursors, neuroepithelial cells and actively proliferating cells. Strikingly, our analysis provides data for identifying at least four other cellular populations that we define as the precursor cells of the growing anlage. Consequently, our findings strengthen the hypothesis that the coelomic epithelium of the anlage is a pluripotent tissue that gives rise to diverse cell types of the regenerating intestinal organ. Moreover, our results provide the initial view into the transcriptomic analysis of cell populations responsible for the amazing regenerative capabilities of echinoderms.

Introduction

Animals exhibit a remarkable diversity in their responses to injury, ranging from basic wound healing to the complete regeneration of lost structures. At one end of the spectrum are species that heal wounds without regenerating the missing part, while at the other end are those capable of recreating structures identical to the original. Despite these differences, all animals possess some level of regenerative ability. This capacity for regeneration can vary not only between different species but also within the same species, depending on the tissue or organ involved.

For decades, scientists have tried to understand these vast differences in regeneration capacities across the animal kingdom. For this they have focused on those species that show amazing regenerative abilities (Tanaka and Reddien, 2011), including coelenterates (hydra) (Vogg et al., 2019), flatworms (planaria) (Reddien, 2018), fish (zebrafish) (Gemberling et al., 2013) and amphibians (Brockes and Kumar, 2002; Roy and Gatien, 2008; Beck et al., 2009; Joven et al. 2019) among others. These studies have uncovered various mechanisms that regeneration-competent species exhibit to regenerate tissues, entire organs, body parts, and in some cases complete bodies.

Key findings include the discovery of essential processes, such as the formation of a blastema – a mass of proliferating cells that plays a crucial role in regenerating the lost structure (Min and Whited, 2023; Poleo et al., 2001; Santos-Ruiz et al., 2002; Seifert and Muneoka, 2018; Stocum, 2004; Zenjari et al., 1996).

Among deuterostomes, echinoderms are considered prime exponents of regenerative capability (Candia-Carnevali et al., 2024). Within this group, holothurians, commonly known as sea cucumbers, exhibit an extraordinary form of regeneration. They can regenerate their internal organs following evisceration, a process in which they expel their viscera in response to stress or predation (Byrne, 2023). This extraordinary ability makes them a valuable model for studying regeneration in complex organisms. The regeneration of the digestive system in holothurians, in particular, has garnered significant interest (Dolmatov, 2021; Mashanov et al., 2014; Medina-Feliciano and García-Arrarás, 2021; Quispe-Parra et al., 2021; Su et al., 2022). Upon evisceration, the holothurian intestine, which constitutes nearly their entire digestive tract, begins to regenerate from the mesentery, a supportive tissue layer where the original intestine was attached (García-Arrarás et al., 1998). Figure 1 and S1 Fig provides a schematic view of the regeneration process and identifies key structures for those not familiar with the holothurian model. A thickening at the injured end of the mesentery, known as an “anlage” or “rudiment,” initiates this process (García-Arrarás et al., 1998). As will be discuss later, this structure is analogous to a blastema but differs in that the cell proliferation mainly occurs in the surrounding epithelium, rather than in the mesenchymal cells, as observed in classical blastemas (Carlson, 2007; García-Arrarás et al., 1998).

Intestinal regeneration in H. glaberrima.

(A) Schematic view of normal and regenerating tissues as viewed in cross sections of normal and regenerating intestines. After evisceration (1hr) the tip of the mesentery is torn but eventually is covered by coelomic epithelium (24hrs). A thickening of the mesentery tip forms the intestinal anlage which grows in size for the next two weeks, initially (3-5 dpe) by cellular dediFerentiation and later by cellular proliferation (7-9 dpe). Eventually, the luminal epithelium is formed from migrating cells of the esophagus and cloacal ends of the digestive tract. (B) DAPI-stained section of 9-dpe anlage and mesentery showing the coelomic epithelium (CE) and connective tissue (CT) layers. (C-D) Sections with fluorescently-labeled phalloidin show the muscle (Mu) labeled in the (C) 9-dpe anlage and in the (D) normal intestine. The normal intestine is made of three distinct layers: mesothelium (Me) (that includes the CE and Mu), CT, and luminal epithelium (not shown). Me and CT are continuous throughout the mesentery and body wall. CE - Coelomic epithelium, CT - Connective tissue, Me - Mesothelium, Mu - Muscle, dpe - days post evisceration. Bar= 10µm

Histologically, the holothurian intestinal anlage forms from dedifferentiated cells within the mesentery, which revert to a more stem-cell-like state before proliferating and migrating to form a new intestinal structure (Candelaria et al., 2006; Mashanov et al., 2005). This dedifferentiation process is crucial for regeneration, involving a spatial and temporal gradient starting at the injury site and extending along the mesentery border (García-Arrarás et al., 2011). The new coelomic epithelial layer that forms around the anlage is distinct from the original mesothelium, shows significant morphological and molecular changes compared to the mesenteric tissue (Candelaria et al., 2006; García-Arrarás et al., 1998; Mashanov et al., 2005). Further examination of holothurian regeneration reveals that most cellular division occurs within the anlage’s coelomic epithelium (García-Arrarás et al., 2011, 1998). These proliferating cells are hypothesized to differentiate into various cell types, including myocytes and neurons, as well as mesenchymal cells, through an epithelial to mesenchymal transition (EMT) (García-Arrarás et al., 2011, 1998). The gene expression profiles during the formation and growth of the anlage suggest extensive reprogramming, leading to a more plastic cell phenotype (Ortiz-Pineda et al., 2009; Quispe-Parra et al., 2021; Rojas-Cartagena et al., 2007).

The process of intestinal regeneration in holothurians raises fundamental questions of regenerative phenomena particularly concerning the identity, origin, and fate of progenitor cells, involved in the process (Alvarado and Tsonis, 2006; Candia-Carnevali et al., 2024). In this context, the role of the anlage, and specifically the mesothelium (also named celothelium) in echinoderms, deserved particular attention (Candia-Carnevali et al., 2024; Smiley, 1994).

This tissue, mainly composed of coelomic epithelia and myocytes, exhibit significant morphological and gene expression changes that are associated with the dedifferentiation process. These dedifferentiated cells form the coelomic epithelium of the anlage and appear to be the principal source of cells for the new intestine. Despite these findings, little is known about the cell composition and dynamics of the anlage nor of its coelomic epithelium.

Single-cell RNA sequencing (scRNA-seq) offers tremendous promise to dissect the cellular contributions of the holothurian intestinal anlage and to identify the specific cells involved in generating a new intestine. It is a powerful tool for dissecting cellular composition and dynamics that has been used in related species to explore regenerating or developing tissues, in mammals (Ayyaz et al., 2019; Beumer and Clevers, 2021; Capdevila et al., 2021; Parigi et al., 2022),axolotl and planaria (Gerber et al., 2018; King et al., 2024; Leigh et al., 2018; Rodgers et al., 2020), and echinoderm embryos (Cocurullo et al., 2023; Paganos et al., 2022a, 2022b, 2021; Satoh et al., 2022; Tominaga et al., 2023). These applications highlight the groundbreaking role of scRNA-seq in advancing our knowledge of the cellular mechanisms in both regenerative and developmental contexts.

In this study, we employ scRNA-seq to analyze the regenerating intestinal anlage of the sea cucumber Holothuria glaberrima, aiming to delineate its constituent cellular populations. We corroborate our findings using hybridization chain reaction fluorescent in situ hybridization (HCR-FISH) to verify the presence and location of specific cell types (Choi et al., 2018). The characterization of the cellular populations and their gene expressions serves to answer various questions such as: What are the cellular precursors? What is the role of the coelomic epithelium? What are the similarities and differences of the anlage and a classical blastema? This research not only advances our understanding of the unique regenerative capabilities of holothurians but also contributes to the broader field of regenerative biology, highlighting the diverse strategies employed by different organisms to restore lost tissues.

Results

Previous work from our laboratory has shown that the rudiment or anlage that forms at the tip of the mesentery plays a pivotal role in the formation of the new intestine (García-Arrarás et al., 2019). This transient mass of cells is thought to give rise to most intestinal cell types, the sole exception being the luminal cells. Therefore, to maximize the characterization of the cells in the regenerative anlage, we chose to perform scRNA-seq in the tissues of 9-day post evisceration (dpe) regenerating animals (García-Arrarás et al., 2011, 1998). At this stage a well-formed anlage consists of epithelial cells surrounding a large area of connective tissue populated with mesenchymal cells (Fig 1A). More importantly, different cell populations undergoing proliferation or differentiation can be found at this stage.

Since some of the cellular processes during regeneration occur in a spatio-temporal gradient along the length of the mesentery, we separately surveyed the anlage and the mesentery tissues accordingly. Thus, for each animal, the anlage was separated from the mesentery, and both tissues were processed independently, for a total of four scRNA-seq runs.

Cell Heterogeneity: Immuno- and cyto-chemical analyses

The strength of the scRNA-seq data depends mainly on the dissociation and isolation of the cell populations from the dissected tissue. Since our focus was on the cells of the regenerating anlage, we devised a dissociation protocol that favored the isolation of the cells within this structure. To determine, at least partially, the cell types in our original dissociation, we performed immuno- and cyto-chemical analyses on the enzyme-dissociated cell suspension that was used for the scRNA-seq. The labeling obtained for each marker is shown in S2 Fig.

We found that, in the dissociated anlage, 5% of the cells were labeled with a spherulocyte (immune cell) marker, 2-7% with phalloidin (a muscle marker), 40% with a mesenchymal marker (KL14-antibody), and a large number of cells (32-60%) with a mesothelial marker (MESO-antibody). Similar populations were found in the mesentery, although in this tissue, 5-10% of cells expressed the neuronal markers, heptapeptide GFSKLYFamide (Diaz-Miranda et al., 1995) and RN1 (Diaz-Balzac et al., 2007). This analysis suggested that most of the cells originate from the regenerate coelomic epithelium.

Different populations were also observed with three different tubulin antibodies (S3 Fig). In the 9-dpe anlage, anti-acetylated alpha tubulin labeled about 7% of the cells, anti-beta tubulin labeled about 70% of the cells, while an anti-alpha tubulin labeled about 80% of the cells. Except for the acetylated alpha tubulin, which labeled about 1% of the cells dissociated from the mesentery versus 7% of those from the anlage, anti-alpha and beta labeling percentages were similar in both mesentery and anlage dissociated cells.

Finally, a cell population labeled with fluorescently-labeled phalloidin accounted for about 7% of the cells in both mesentery and anlage. These cells, however, did not correspond to the elongated muscle cells of the mesentery. Instead, they were rounded cells with labeling found in the cytoplasm surrounding one side of the nuclei.

Two cell populations from the mesentery were absent or greatly underrepresented in the scRNA-seq. Firstly, the muscle cells of the mesentery, due to their non-dissociation by the protocol used, were not found in the dissociated cell suspension. Their elongated morphology would have further complicated their passage via cell separation system for sequencing. Secondly, the majority of neurons from the neuronal network associated with the mesentery (Nieves-Ríos et al., 2020) could not be dissociated. We did observe structures that resembled a tangled mass of cells immunoreactive to some of our neuronal markers. This suggests that the mesentery nervous component, being unable to be isolated as single cells, was not sequenced.

In summary, the immuno- and cytochemical results show that the dissociated cell populations sequenced correspond to cellular phenotypes that have been previously described within the regenerating anlage. The abundance of these cells in the sequenced samples corresponds to the ease of their dissociation by trypsin. Thus, dedifferentiated cells of the mesentery and anlage epithelium (which are loosely connected to each other) and those of the connective tissue are probably over-represented compared to differentiated cell types.

Cell populations defined by scRNA-seq

Analysis of scRNA-seq data resulted in a total of 3,844 cells, with 2,392 originating from the two anlage samples and 1,452 from the two mesentery samples (S4 Fig). Upon dataset integration and graph-based clustering, we identified 13 clusters, each thought to represent singular cell types or cell states in the regenerating intestine (Fig 2A & S5 Fig). The percentage of cells that form each cluster differs from cluster to cluster, ranging from 21% (cluster 0) to 1% (cluster 12) of the total cells. Nonetheless, each cluster consists of cells from both the mesentery and anlage samples (Fig. 2B & C). The number of clusters did not change dramatically under various parameters (resolution, dimensionality, and number of variable features), constantly around 12-15 clusters. Moreover, except for C3 and C4, the clusters were supported by the clustering significance analysis performed with scSHC (Grabski et al., 2023), a model-based hypothesis testing method for scRNA-seq that evaluates the probability of each individual cluster being unique (S5 Fig). The uniqueness of C3 and C4 is suggested by additional analyses as will be addressed below.

Overview of single cell RNA sequencing of regenerating intestinal tissue of H. glaberrima.

(A) UMAP plot of population identities determined through unsupervised clustering of 9-day regenerating mesentery and anlage tissues. (B) UMAP projections of cluster cells separated by tissue of origin. (C) Percentage of cells per cluster based on their tissue of origin.

Each identified cluster exhibited a distinctive gene expression profile relative to cells in other clusters (S6 Fig). This shows the relative expression of the top gene for each cluster based on two factors: (1) difference in percentage of representation and (2) log2 fold-change (log2FC) against all other clusters. Interestingly, each cluster shows dramatic differential expression values (> 2 log2FC) and differences in representation percentages over 50%, except for C0 through C3, with differences in representation around 30%.

Prior to characterizing each of the 13 cell clusters, we sought to understand what, in a broad view, appeared to be a segregation of ∼90% of the cells into two distinct supra-clusters. One of them encompassing 7 clusters (C0, C1, C3, C4, C5, C8, C9) that corresponded to 69.6% of all cells and the other encompassing two clusters (C2 and C7) that corresponded to 19.1% of cells. The remaining 11.3% of cells were distributed in 4 distinct isolated clusters (C6, C10, C11, and C12). As stated earlier, all clusters have representation from mesentery and anlage tissues (Fig 2B-C), thus excluding the possibility that the two supra-clusters represented mesentery versus anlage cells.

Mesenchymal versus Epithelial Clusters

These two supra-clusters are of interest as they appear to represent the two main cell types found in the regenerating intestine: coelomic epithelial cells and mesenchymal cells (Fig 2A). Comparison between these two supra-clusters showed distinct expression profiles that allowed us to characterize their cell types (Fig 3 and S6 Fig).

Cluster characterization by gene expression.

(A) Top-expressed genes or (B) genes corresponding to transcription factors and intercellular signaling molecules are identified in the 13 cell clusters. Clusters are classified by their corresponding cell type, where blue corresponds to cells of the coelomic epithelium, red to those in the mesenchyme and green to coelomocytes. Color intensity shows the expression level of each gene in log2fold-change (log2FC) values. Dot size corresponds to percentage of representation of the gene in the respective cluster compared to all others. Gene identifiers starting with “g” correspond to uncharacterized gene models of H. glaberrima. (C) UMAP plot of clusters colored by cell type.

For instance, the supra-cluster composed of C2 and C7, showed ERG (transcriptional regulator ETS-related gene) as the top expressed gene, an oncogene that is associated with mesenchymal cells in other echinoderms (Meyer et al., 2023; Tominaga et al., 2023). While this gene has a strong expression in C2 and C7, it also appears to be expressed in C6 and C10. However, it is important to highlight that ERG as a transcription factor has been shown to have roles in additional processes including inflammation and apoptosis .

The top marker gene in the other supra-cluster is AHNK, known as neuroblast differentiation-associated protein AHNAK. Reports have shown AHNK to have a role in calcium regulation, cellular migration, and carcinogenic transformation of colon epithelial cells (Dumitru et al., 2013). Furthermore, this gene is overexpressed in regenerating rat muscle compared to normal muscle (Huang et al., 2007). However, the localized expression of WNT9 in the C1, C3, C5, and C8 of this supra-cluster more clearly favors its classification as a marker for coelomic epithelial cell types (Fig 3B and S7 Fig). A previous study from our laboratory using in situ hybridization, details the expression of WNT9 during intestinal regeneration in H. glaberrima, where it was shown to be localized to the coelomic epithelium of the anlage and adjacent mesentery (Mashanov et al., 2012). In addition, correlating with what was shown by the in situ hybridization results, the population of cells that differentially expresses WNT9 makes 15% of those in the mesentery. However, it is close to 30% of the cells in the anlage. Additional analyses, discussed below, further strengthens the coelomic epithelium identity of cells in this supra-cluster.

In summary, results show 13 individual cell clusters with distinct expression profiles in the regenerative intestinal tissue with most of them corresponding to either mesenchymal or coelomic epithelia cells. The rest of the populations show top expressed genes that are immune-related suggesting that these must be coelomocyte populations such as those that have been previously associated with both normal and regenerating intestinal tissues and that were also detected by immune and histochemistry in the cell samples used for the scRNA-seq (S2 Fig) (Garcia-Arraras et al. 2006; Ramirez-Gomez et al. 2010).

Cluster Identities

Rather than considering single genes, the uniqueness of each cluster can be assessed in terms of their transcriptomic profile (Fig 3A) and by the expression of transcription factors and intercellular signaling molecules (Fig 3B). As shown in Fig 3A-B, each cluster has a unique transcriptomic profile. Nonetheless, some clusters show their close relationship based on shared gene expression or representation. A notable example is clusters C0, C1, C3, and C4 of the epithelial layer populations, which share many top genes, albeit at different expression levels.

Similarly, C8 and C9 have low expression of genes expressed by C0, C1, C3 and C4 but have expression of other genes that are not expressed by any other cluster. This is also true for the mesenchymal clusters, where these two clusters have overlap of some genes, that are not expressed by other clusters. Therefore, these results show the interaction of these clusters and their transcriptomic relationship depending on where they are localized within the regenerating intestinal tissue.

Identifying the top genes expressed by each cluster also provides essential, notwithstanding limited, information for their complete identification. To further characterize each cluster, we have used other analyses or performed additional experiments, including (1) multiple gene expression patterns, (2) enriched ontology, (3) HCR-FISH and (4) pseudo-trajectory. Initially, we describe in depth the potential identity of these clusters based on their expression pattern and enriched ontology, and we will end with their potential interactions.

An in-depth analysis of the various clusters and their possible relation to cell populations

Coelomocyte populations

Coelomocytes are specialized cells found in the coelomic fluid and within organs of echinoderms. These cells have been associated with immunological roles including pathogen recognition, encapsulation, phagocytosis, debris removal, cytokine production and secretion among others (Barela Hudgell et al., 2022; Courtney Smith et al., 2018; García-Arrarás et al., 2006; Smith et al., 1995), In holothurians, these cells have been shown to be present at injury sites and in the mesentery and regeneration anlage (García-Arrarás et al., 2006). Coelomocytes can be subdivided into different populations by using morphological, physiological, and molecular characteristics (Ramírez-Gómez et al., 2010).

Many of the coelomocyte characteristics correlate with the top differentially expressed genes of cell populations within our data. These immune-like clusters (C6, C10, C11, and C12) have high expression of genes related to the immune system that are not shared with any other cluster. For example, C6 embodies a distinct cell population that represents a substantial number of cells (6% of all cells, 8% of the mesentery and 4% of the cells in the anlage).

These cells are the only population expressing FBCD1 (fibrinogen C domain-containing protein 1). Cells in C6 also express other genes such as various tyrosine protein phosphate receptors, integrin alpha-8 (ITA8), platelet glycoprotein V (GPV), leucine rich repeats and immunoglobulin-like domains protein 2 (LRIG2) (Fig 3). The immune identity of C6 is also supported by the resulting gene set enrichment of gene ontology (gseGO) terms related to immune responses, such as ubiquitin-dependent ERAD pathway, innate immune response activating cell surface receptor signaling pathway, respond to endoplasmic reticulum stress, phagocytosis and many more related to defense mechanisms (S8 Fig). Similarly, C10 and C12 transcriptomic profiles suggest they correspond to immune-like populations (Fig 3). Specifically, C10 shows various uncharacterized genes along with HCK (tyrosine-protein kinase HCK), DMBT1 (deleted in malignant brain tumors 1), ALS (insulin-like growth factor binding protein complex acid labile subunit), GPV, and IRF8 (interferon regulatory factor 8), FER (tyrosine-protein kinase Fer). The top GO enriched terms of this cluster support its involvement in immune process, some of these being: lipase activity regulation, Fc receptor mediated stimulatory signaling pathway, cellular pigmentation, B cell activation involved in immune response, and Fc receptor signaling pathway (S8 Fig). The transcripts expressed by C12 show a more complex profile, where most of the top genes are uncharacterized. Nonetheless, among the annotated genes are BAR3 (Balbiani ring protein 3), LYS1 (lysozyme 1), TRPM3 (transient receptor cation channel subfamily M member 3), PGCA (aggrecan core protein) and MRC1 (macrophage mannose receptor 1). The top genes of C11 include a great number of immune genes such as MUC5A (mucin-5AC), FCGBP (IgGFc-binding protein), SSPO (SCO-spondin), FCN1A (ficolin-1-A), MUC5B (mucin-5B), and TIE1 (tyrosine-protein kinase receptor Tie-1). However, the top genes of this cluster also include several genes involved in neuronal activity, which is evident in the top enriched GO terms of this cluster that include regulation of postsynaptic membrane potential, excitatory postsynaptic potential, chemical synaptic transmission, endoplasmic reticulum to Golgi vesicle-mediated transport, adult behavior, and regulation of neurotransmitter levels (S8 Fig).

To partially confirm the prediction that cells from C6, C10, C11 and C12 corresponded to coelomocyte populations, we used HCR-FISH to identify the cell types expressing the top gene in two of the clusters. We focused on the expression of FBCD1 and BAR3, the genes that are the most represented by cells of C6 and C12, respectively. For each probe, in situ hybridization identified a distinct cell type in both regenerating and non-regenerating tissues (Fig 4, S9 Fig & S10 Fig). The FBC1-expressing cells showed round or oval morphologies with a central round nucleus. In some cases, short extensions could be observed. The cells were heterogeneously distributed in all tissues, including the nervous system and the body wall, and could be found associated with either epithelial tissues or with the extracellular matrix (ECM) in the normal intestine, the intestinal anlage and in the mesentery of normal or regenerating animals (Fig 4A & 4C). The BAR3-expressing cells were also distinct, isolated cells found in different tissues of the normal and regenerating animals (Fig 4B & 4D). Their numbers were not as high as those of the FBC1-labeled cells and their labeling was more punctuated within the cytoplasm. In the regenerating tissues they were mostly associated with the mesentery. The widespread distribution of both cell types hinted at a cell function consistent with patrolling the body to detect and respond to potential threats such as injury or bacterial invasion.

Expression profile of coelomocyte cell types.

HCR-FISH for (A, C) FBCD1 or (B, D) BAR3 in holothurian (A, B) normal or (C, D) regenerating tissues. Cells (arrowheads) expressing FBCD1 mRNA in the connective tissue (CT) layer of (A) normal intestine and (C) anlage. Cells (arrowheads) expressing BAR3 mRNA in the connective tissue of (B) normal (non-eviscerated) animal and (D) the coelomic epithelium (CE) of the anlage. Insets provide the approximate localization in the connective tissue of the normal intestine (top) and the anlage (bottom). Cyan - DAPI, Magenta - HCR-FISH. CE - Coelomic epithelium, CT - Connective tissue, Mu - Muscle. Bar= 10 µm.

Mesenchymal cell populations

Mesenchymal cells of the intestinal anlage are yet to be well studied. They are known to be less proliferative than those in the overlying epithelium (García-Arrarás et al., 2011, 1998) and involved in ECM remodeling (Quiñones et al., 2002). Some of the mesenchymal cells are thought to migrate from the connective tissue of the mesentery to the connective tissue in the anlage (Cabrera-Serrano and García-Arrarás, 2004), while others have been shown to originate via EMT from the overlying epithelium (García-Arrarás et al., 2011). Some of these cells will form a mesenchymal cellular layer associated with the luminal epithelial cells as the lumen forms (García-Arrarás et al., 2011). As explained earlier, we propose that C2 and C7, form a separate supra-cluster, clearly identified by ERG expression. These two clusters also share many marker genes, including multiple ECM genes (Fig 3). For instance, highest expressed gene of C2 is TIMP3 (metalloproteinase inhibitor 3), followed by NPNT (nephronectin), which has been reported to be an integrin ligand during kidney development (Sun et al., 2018). Additionally, this cluster has overexpression of ECM2 (extracellular matrix protein 2), KLKB1 (plasma kallikrein), MMP14 (matrix metalloproteinase 14), MMP24 (matrix metalloproteinase 24), HMCN1 (hemicentin-1), and ITA8 (integrin alpha-8), all of which are explicitly related to extracellular matrix component (Bokel and Brown, 2002; Dolmatov and Nizhnichenko, 2023; Dong et al., 2006; Stamenkovic, 2003; Volkert et al., 2014). Many of these genes are also highly represented in C7. However, here we also find ITIH2 (inter-alpha-trypsin inhibitor heavy chain H2), DUOX1 (dual oxidase 1), SVEP1 (sushi, von Willebrand factor type A, EGF and pentraxin domain-containing protein 1), SEPP1 (selenoprotein P), and KLH20 (Kelch-like protein 20) suggesting that cells have gained some specialization and are more advanced in their differentiation when compared to those of C2. Along with this, when analyzing the gseGO, for C2 we obtain GO terms of numerous ECM processes, such as cell adhesion mediated by integrin, integrin-mediated signaling pathway, and regulation of cell-substrate junction assembly (S8 Fig). Similarly, C7 has an enrichment of regulation of cell-matrix adhesion, regulation of cell-substrate junction organization, and substrate adhesion-dependent cell spreading. neuroblast proliferation, among others (S8 Fig).

To confirm our prediction that cells in C2 and C7 were those present within the mesenchyme of the connective tissue, we chose to localize the expression of HMCN1 mRNA. The mRNA for this protein, known to code for an extracellular matrix protein (Dolmatov and Nizhnichenko, 2023; Lindsay-Mosher et al., 2020; Welcker et al., 2021), is present in both cell clusters. HCR-FISH showed that cells expressing the HMCN1 mRNA were present in the normal intestine, the regenerating mesentery and the anlage (Fig 5). In the normal intestine and regenerating mesentery, labeled cells were present in the connective tissue (Fig 5A-B). In both tissues, cells were somewhat distanced from each other and had an irregular morphology with strong punctate labeling throughout the cytoplasm. A weaker labeling was observed in cells of the anlage, which were more densely packed and adjacent to the coelomic epithelium (Fig 5C), suggesting that they correspond to cells undergoing EMT on their way to differentiate into ECM-producing mesenchymal cells. A very similar pattern of expression was observed with HCR-FISH for ERG (not shown).

Expression profile of mesenchymal cell types.

Cells (arrowheads) expressing HMCNT1 mRNA in the connective tissue (CT) layer of (A) normal intestine, (B) regenerating mesentery, and (C) intestinal anlage of H. glaberrima. In all tissues, the cells are found within the connective tissue (CT) layer. Notice in C that no expression is found in the cells of the coelomic epithelium (CE). Insets provide the approximate localization of the cells in the adjacent photos. Cyan - DAPI, Magenta - HCR-FISH. CE - Coelomic epithelium, CT - Connective tissue, Mu - Muscle. Bar= 10µm

Coelomic epithelia/mesothelial cell populations

The mesothelial layer of the intestine and mesentery is formed by cells (coelomic epithelium or peritoneocytes together with myocytes and neurons) in contact with the coelomic fluid. The coelomic epithelium of the regenerating tissues differs in morphology and gene expression (Mashanov et al.,2005, 2017, 2015, 2012, 2010) to the mesothelium that normally surrounds the organ. This coelomic epithelium which is present mainly in the anlage and in areas of the adjacent mesentery, is made of dedifferentiated cells and is responsible for most of the cell division that takes place in the regenerating intestine (García-Arrarás et al., 2011). The analysis of the scRNA-seq data, in view of our knowledge of the ongoing events in the 9-dpe regenerating organ, strongly suggest that the 7 clusters within the major supra-cluster represent the cells in the coelomic epithelium of the mesentery and the anlage. Here is our analysis:

C8 represents the proliferating cells

These cells mainly found within the anlage coelomic epithelia express proliferation markers such as PLK1 (serine/threonine-protein kinase PLK1), SMC2 (structural maintenance of chromosomes protein 2), PRI2 (PRIM2 - DNA primase large subunit), PCNA (proliferating cell nuclear antigen), and CDK1 (cyclin-dependent kinase 1) (Locard-Paulet et al., 2022). In addition, this cluster has high expression of TOP2A (DNA topoisomerase 2-beta), a gene that has also been seen to be overexpressed in proliferating basal cells of the human gastrointestinal epithelia (Busslinger et al., 2021). Other marker genes related to cell mitotic activity found here include CENPE (centromere-associated protein E), SMC4 (structural maintenance of chromosomes protein 4), KI67 (proliferation marker protein Ki-67), and CCNB3 (G2/mitotic-specific cyclin-B3) (Fig 3). Furthermore, this cluster expresses specific transcription factors such as E2F3, MCM10 (protein MCM10 homolog), PAF15 (PCNA-associated factor) and BRCA1 (breast cancer type 1 susceptibility protein) that are also associated with control of cell division (Humbert et al., 2000; Lõoke et al., 2017; Xie et al., 2014). The gseGO terms also confirm its proliferative identity with GO terms related to chromosome separation, condensation, mitotic cytokinesis, and regulation of cell cycle checkpoint (S8 Fig). Moreover, the dividing cell population is higher in the anlage samples (4%) than in the mesentery samples (2%), which is in accordance with what we have observed in regenerating animals, that while cell division does take place in the mesentery, more cells are proliferating in the epithelial layer of the intestinal anlage (Bello et al., 2020; García-Arrarás et al., 2011, 1998).

To verify the epithelial nature of this cell cluster we performed HCR-FISH for the KI67 mRNA. Multiple cells of the coelomic epithelium of the anlage were found to express the gene, as determined by a punctate labeling found throughout the cell body (Fig 6A-B). Some cells in the mesentery coelomic epithelium were also labeled but their number decreased in areas closer to the body wall. Similarly, few cells were labeled in the mesothelium of the normal intestine.

Expression profile of proliferating cells.

Cells (arrowheads) expressing Ki67 mRNA in the coelomic epithelium (CE) layer of regenerating (A) mesentery and (B) anlage of H. glaberrima. Most of the labeled cells are within the CE tissue layer. (C) BrdU-labeled cells (arrowheads) are also mainly found in the CE layer. Insets provide the approximate localization of the cells in the adjacent photos. Cyan - DAPI, Magenta - HCR-FISH. CE -Coelomic epithelium, CT- Connective tissue. Bar = 10µm

Technical aspects of the protocol do not allow for double labeling of BrdU and HCR-FISH to be performed. (The HCl step that is needed in BrdU labeling to allow the antibody access to the BrdU-labelled DNA seems to interfere with the HCR-FISH labeling). Nonetheless, we could show that, as previously documented (Garica-Arraras et al., 1998, 2011), BrdU-labeled are found in the coelomic epithelium of the anlage (Fig 6C). The number of BrdU cells is much smaller than that of KI67 HCR-FISH labeled cells. This is expected since BrdU is only incorporated into the DNA in cells undergoing the S-phase while KI67 is expressed during the whole duration of the mitotic cycle (Kee et al. 2002).

C5 represents muscle precursors

Enteric muscle precursors are known to originate from the coelomic epithelium during the second week of regeneration (Murray and García-Arrarás, 2004). This population can be recognized in our data by the expression of muscle-specific markers present in C5, such as TITIN, MYL1 (myosin light chain 1/3), MYH7 (myosin 7), ACTG (actin, cytoplasmic 2), and CNN3 (calponin-3) (Fig 3). Considering that this cell population is potentially a population of the coelomic epithelium actively undergoing differentiation toward muscle phenotype, they also share gene expression with other epithelial cell clusters (C0, C1, C3, C4), albeit at a lower fold-change.

Similarly, this cluster shows a specific expression of transcription factors associated to muscle cells, namely FXL16 (F-box/LRR-repeat protein 16) and SCRT2 (transcriptional repressor scratch 2). Results of GO terms of this cluster also demonstrate enriched terms related to muscle tissue growth, such as muscle tissue morphogenesis, muscle development, myofibril assembly, and sarcomere organization (S8 Fig). The focus on development and morphogenesis is to be expected, considering that these are still undergoing differentiation toward a muscle phenotype. HCR-FISH of MYH7 corroborates the muscle phenotype of the cells in this cluster (Fig 7). In control intestinal tissues the labeling is specific to the muscle cell layer (Fig 7 and S11 Fig, also see Fig. 1). In regenerating tissues the labeling is observed in some cells in the basal part of the ceolomic epithelium (Fig 6), the same regions as where myoblasts or muscle cells were previously identified (Murray and García-Arrarás, 2004) (Fig 7). An additional experiment was performed to double label the tissue with MYH7 HCR-FISH and an antibody that labels the muscle fibers. The overlay of these markers showed the expression of the myosin mRNA in the normal intestine muscle cells (Fig 7 D & 7E) and in cells of the regenerating tissues that have begun to form the actin-myosin contractile apparatus (Fig 7F & 7G).

Myosin mRNA expression in regenerating and mature muscle cells.

(A-C) Cells expressing MYH7 mRNA (arrowheads) in the (A) normal intestine, (B) regenerating mesentery, and (C) basal area of the anlage coelomic epithelium (CE). Arrowheads point to expression in cells of regenerating tissues. Insets provide the approximate localization of the cells in the adjacent photo. (D-G) Double labeling of (D, F) MYH7, and (E, G) muscle-specific antibody (HgM2), in (D, E) normal intestine, and (F, G) regenerating mesentery, showing the co-expression of both markers in the same cells. Note the two cells (arrowheads in F-G) that express both the MYH7 marker and the muscle-specific antibody (HgM2), representing cells initiating their diFerentiation toward enteric muscle cells. Cyan - DAPI, Magenta - HCR-FISH, Yellow - muscle antibody (HgM2). CE - Coelomic epithelium, CT - Connective tissue. Mu - Muscle. Bars = 10μm.

C9 represents the neuroepithelial cells

Cells in C9, represent another population of specialized cells that can be associated to cells previously described in the intestinal anlage. This small number (3%) of cells most likely corresponds to neuroepithelial cells that will eventually give rise to neurons. These cells express neuroepithelial or neuronal genes such as neurotrypsin (NETR), potassium gated-voltage channels, PRD10 (PRDM10 - PR domain zinc finger protein 10), ELAV2 and STA10 (STARD10 - START domain-containing protein 10) (Fig 3). The latter is a protein that we have characterized as being expressed by enteric neurons and nerve bundles (Rosado-Olivieri et al., 2017). The holothurian STA10 (STARD10) is recognized by our monoclonal antibody RN1. This antibody has been used to detect enteric neurons as they begin to differentiate in the coelomic epithelium during the second week of regeneration (Tossas et al., 2014). Among its associated GO terms are positive regulation of ion transmembrane transporter activity, cyclic nucleotide metabolic process, regulation of muscle contraction, regulation of membrane potential and positive regulation of hormone secretion (S8 Fig). Moreover, there is also a great representation of processes involved in development, differentiation, and growth of nerve cells, all of which together would be expected of a neuroepithelial layer. To verify the neuroepithelial nature of the cells we performed HCR-FISH for the STA10 mRNA in the tissue and show that some cells in the regenerating mesentery and anlage coelomic epithelium express the gene (Fig 8). Moreover, double labeling with the HCR-FISH against STA10 and the RN1 antibody show cells and fibers that express both the mRNA and the protein product (Fig 8) are forming fiber extensions typical of differentiating neurons (Fig 8C-E).

STARD-10 expression in differentiating neuroepithelial cells of the coelomic epithelium.

(A) Cells (arrowhead) expressing STA10 mRNA in the coelomic epithelium (CE) of the regenerating mesentery. (B) Double labeling with STA10 HCR-FISH and STARD-10 antibody (RN1) shows the presence of a cell (arrowhead) in the CE of the regenerating mesentery. (C-E) Fibers (arrow) in the anlage, also express both the (C) STA10 mRNA and (D) the protein, as shown in (E) the corresponding overlay. Insets provide the approximate localization of the cells or fibers in the adjacent photos. Cyan - DAPI, Magenta - HCR-FISH, Yellow-STARD-10 antibody (RN1). CE - Coelomic epithelium, CT - Connective tissue. Bar- B = 10μm; A, C-E = 5 μm.

The characterization of these 3 clusters (C5, C8, C9), that represent cells undergoing differentiation or proliferation, leaves a group of 4 clusters (C0, C1, C3, and C4) that show some overlap in expressed genes and at the same time share some gene expression with some of the previously described clusters. Nonetheless, as seen in Fig 3A-B, the cells in these four clusters still have high expression and representation of specific transcripts.

C4 represents the intestinal coelomic epithelial cells

The gene expression profile of cells in C4 sets them slightly apart from the other three clusters (C0, C1, C3). It identifies cells that are more advanced in their development toward a particular phenotype. C4 shows high expression of genes such KCNQ5 (potassium voltage-gated channel subfamily KQT member 5), SC6A9 (sodium and chloride dependent glycine transporter 1), EFNB2 (Ephrin-B2), and UNC5C (Netrin receptor UNC5C) (Fig 3). Other than these, C4 also shows high expression of genes that are related to cell-cell interactions and ECM molecules such as LAMA2 (laminin subunit alpha-2),

MEGF6 (multiple epidermal growth factor-like domains protein 6), FMN1 (formin-1), NPHN (nephrin). This cluster, distinct from others, shows enrichment of GO terms related to more advanced stages of development, such as morphogenesis of epithelium, sensory perception, regulation of calcium ion transmembrane transport, among others (S8 Fig). Noteworthy, cells from this cluster correspond mostly to cell from the mesentery samples (67%) rather than from the anlage (37%). Additionally, from all the mesentery cells, about 11% are part of this cluster, while only 4% of the anlage cells are represented here. HCR-FISH of UNC5C (Netrin Receptor), a gene differentially expressed in the cells of this cluster provided a surprising result. The expression of this mRNA was observed as an intense labeling in cells of the normal intestine mesothelium (Fig 9A), In the regenerating tissue, the coelomic epithelium of the mesentery is labeled (Fig 9B-C) and some labeling is observed in the coelomic epithelium of the anlage (Fig 9D). This pattern of labeling clearly shows the cells to be part of the coelomic epithelium and strongly suggests that they correspond to cells that are in a differentiation pathway to become part of the coelomic epithelia (possibly the peritoneocytes) of the regenerated organ. This conclusion is strengthened by the pseudotime analyses presented in the following section.

Expression profile of coelomic cell populations in the regenerating intestine anlage and mesentery.

(A) HCR-FISH for UNC5C mRNA labeled most if not all cells in the coelomic epithelia (CE) of the normal intestine. Strong labeling is also observed in various regions of (B, C) the CE of the regenerating mesentery, while a few cells (arrowhead) are found in (D) the CE of the anlage (arrowhead). Insets provide the approximate localization of the cells in the adjacent photos. Cyan - DAPI, Magenta - HCR- FISH. CE - Coelomic epithelium, CT - Connective tissue. Bar = 10µm.

C0, C1, and C3 represent differentiation stages of coelomic epithelial cells

The three remaining clusters to be analyzed are C0, C1, and C3. These three clusters share many of their top representative genes, which are associated with developmental, regenerative, or oncogenic processes (Bradford et al., 2009; Dunn et al., 2006; Lu et al., 2015; Nishimoto and Nishida, 2007; Oike et al., 2004; Zhao et al., 2008). These include, for C0: TRFM (melanotransferrin), TIMP4 (metalloproteinase inhibitor 4), and DMBT1, for C1: FGF13 (fibroblast growth factor 13), TGFB3 (transforming growth factor beta 3), HS90A (heat shock protein HSP 90-alpha), and ANGL1 (angiopoietin-related protein 1) (Fig 3), and for C3: SEM5B (semaphoring-5B), LRIG3 (leucine-rich repeats and immunoglobulin-like domains protein 3), TUTLB (protein turtle homolog B), and NET1 (netrin-1). Moreover, C0 and C1 share an over-representation of GO enriched terms that highlights biological processes related to ribosomal activity such as cytoplasmic translation, and ribosome assembly, biogenesis, and assembly (S8 Fig). In addition to these, C3 also showed enrichment of processes involved in the development of tubular lumen-containing structures such as mesonephric and ureteric ducts, differentiation and regulation of cell growth, and negative regulation of axogenesis (S8 Fig).

C1 and C3 are also closely related in their localization, being overwhelmingly associated with the anlage. These two clusters are mostly composed of cells from the anlage tissue, where we expect to see the precursor cells that will give rise to specialized cells of the organ (Fig 1). Precisely, C1 and C3 cells together correspond to 37% of all anlage cells, compared to 14.6% of all mesentery cells. In contrast, cells from C0 correspond to 27% and 18% of the mesentery and anlage cells, respectively.

To obtain insight into these cell clusters, we performed HCR-FISH for two different mRNAs; NET1 (Netrin), a chemotropic protein highly represented in C1 and C3, and TRFM (Melanotransferin) the top represented gene in C0. Both in situ hybridization experiments labeled cells in the coelomic epithelium of the regenerating intestine, supporting our contention that the large supra-cluster represents the coelomic epithelium layer (Fig 10). However, their spatial pattern of expression was unpredictably different. While NET1 was highly expressed in most of the coelomic epithelial cells of the anlage, little expression was found in the regenerating mesentery or in the coelomic epithelium of the normal intestine (Fig 10).TRFM, in contrast, was highly expressed in the coelomic epithelium of the normal intestine and poorly expressed in the intestinal anlage (Fig 10). In the regenerating mesentery, a gradient in expression of the TRFM is observed, where high levels of expression were found in the coelomic epithelium close to the body wall and diminished as one approached the anlage. Thus, the HCR-FISH results show that C0, C1, and C3 correspond to cells of the coelomic epithelium, but strongly suggest that C0 differs from C1 and C3 both in their gene expression profile and in the localization where they are found, both in the regenerating and in the normal intestine.

HCR-FISH for TRFM and NET1 shows the differential gene expression of the anlage versus mesentery coelomic epithelia.

TRFM mRNA (left column) is expressed by (A) cells (arrowheads) of the coelomic epithelium (CE) in the normal intestine. In regenerating tissues, there is a high expression in (C) CE cells (arrowheads) close to the body wall. This expression decreases toward the anlage, with some expression in (E) CE cells of the mid part of the mesentery and very little if any in (G) the CE of the anlage. In contrast, NET1 (right column) is expressed in fewer CE cells (arrowheads) of the (B) normal intestine, and (D, F) mesentery, but its expression is extremely high in most of (H) the CE cells of the anlage. Insets provide the approximate localization of the cells in the adjacent photos. Cyan - DAPI, Magenta - HCR-FISH. CE - Coelomic epithelium, CT - Connective tissue. Ba r= 10µm.

Finally, it is essential to highlight the many signaling, or growth factors expressed by the coelomic epithelial clusters, in particular C0, C1 and C3 (Fig 3). These include: Wnt, Hox, semaphorin, FGFs, TGF-beta, netrin, insulin-like growth factor (IGF), growth/differentiation factors (GDF), and angiopoietin related proteins (ANGL) (e.g. WNT9, SEM5B, FGF13, NKx3.2, TGFB3, HOX9, IGF1, GDF8, ANGL1, and FOXF1) (Fig 3). This is important in view that the epithelium of the vertebrate blastema is characterized by its chemical modulation of the underlying mesenchyme, as will be discussed later. Likewise, other genes that serve as markers of specific cell types or cellular stages were also identified, including PIWL1 (piwi-like protein 1) in C1, YAP1 in C3 and C4, HES1 (transcription factor HES-1) in C1 and C3, and PRRX1 (paired mesoderm homeobox protein) in C0 and C4 (Fig 3 and S12 Fig).

In summary, of the 13 clusters identified, our data strongly suggest that four of them (C9, C10, C11, and C12) correspond to coelomocytes or immune cells, two of them (C2 and C7) correspond to cells with a mesenchymal phenotype and the remaining 7 to cells of the coelomic epithelia. Of these seven, C5 corresponds to differentiating muscle, C9 to differentiating neuroepithelium, C4 to differentiating coelomic epithelia and C8 to proliferating cells. C1 and C3 represent most of the cells found in the coelomic epithelium of the anlage. In contrast, C0 represents a coelomic epithelial phenotype more closely associated with the mesentery than the intestine.

Trajectory Analysis: What populations are driving cell specification?

Among the many mysteries of the holothurian intestinal regeneration process is the identification of the precursor cells. In simple terms, what are the cells from which all nascent cells derive from? We performed a trajectory analysis of the data to provide some insights into this issue. This type of analysis is usually performed with samples at different stages or time points. However, we considered it feasible to conduct this analysis because in the 9-dpe regenerating anlage/mesentery we find cells at various stages of differentiation. These cells could provide crucial information on how the cell populations are associated with each other. To address this, we initially employed RNA velocity analysis. This method describes the temporal dynamics of gene expression based on the relative abundances of spliced and un-spliced mRNA across cell populations.

Our initial velocity analysis on all the clusters and samples (Fig 11A-B), provided three main results. First, the direction of arrows in our UMAP shows them flowing towards C5, C9, and C4. These arrows do not point towards any other cluster; thus, they are terminal arrows. Second, while arrows from C8 are not terminal, they are directed towards C1. This suggests that cells in C1 provide cells for the growth of the anlage via proliferation. Thus, these results support C5, C9 and C4 as terminal cell clusters which we have described as muscle, neuroepithelial, and the nascent coelomic epithelium cells, respectively. Third, velocity embedding shows shorter arrows that point from clusters 0 and 1 towards terminal populations previously described. Therefore, cells of C0 and C1 are not undergoing significant transcriptional changes. The RNA velocity results of the rest of the clusters are less interesting as they do not show directions towards any other clusters, mainly because of their individuality within the UMAP. However, it is interesting that arrows of the mesenchymal cell populations show distinct directions and lengths. Based on the results, it seems that portions of both mesenchymal clusters (C2 and C7) have gone or are undergoing more extensive differentiation changes.

Trajectory analysis of cell populations from the regenerating intestinal tissue.

(A) UMAP plot of all identified clusters. (B) RNA velocity embedded in UMAP of all main clusters. (C) UMAP of re-clustering of cells from C0, C1, C3, C4, C5, and C9. (D) RNA velocity analysis results from the subset from panel (C). (E) Jitter plot of Slingshot pseudotime of cells from C0, C1, C3, C4, C5 and C9. Pseudotime resulted in two lineages one containing C5 and the other C 9. (F) UMAP of re-clustered anlage cells corresponding to C0, C1, C3, and C4(G) RNA velocity results from the cluster results from panel F. (H) UMAP of panel F plot overlayed with pseudotime results of Slingshot. Color represent pseudotime values from 0 (blue) to 60 (yellow). (I) Jitter plot showing the Slingshot pseudotiume of cells from each cell cluster.

We then re-clustered C0, C1, and C3 cells, along with the differentiating cell populations (C4, C5, and C9) to better understand their relationship (Fig 11C). The velocity assessment of these newly clustered populations resulted in a similar pattern (Fig 11D). Here, we can see that the differentiating cells (C4, C5, and C9) have long arrows suggesting that these cells are going through an advanced stage of transcriptional change compared to others (Anderson et al., 2020; La Manno et al., 2018). It is clear from these results that C1 future states are cells of C3 and, to some extent, C4. Moreover, it seems that C0 has a closer relationship to cells of C5 and C9 and that some cells of these clusters could potentially differentiate into cells of C4 (the coelomic epithelium cells). In this case, to complement and confirm our RNA velocity interpretation, we also performed a pseudotime analysis using Slingshot, which relies on the expression data of each cluster. This analysis showed C1 in an earlier pseudotime than C3, C4 and C0 in the resulting two lineages (Fig 11D). The resulting lineages differed by the terminal clusters, one containing C5 (muscle) and the other C9 (neuroepithelial). Thus, it supports what we have already visualized on the RNA velocity embeddings.

The results described so far show that C0, C1 and C4 are cells in distinct differentiation states, but we wanted to have a clearer view of the cell clusters that potentially have an essential role in the regeneration process. For this, we made another subset of cells that corresponded to C0, C1, C3 and C4, but uniquely from cells of the anlage (Fig 11F). The rationale was that cells from the mesentery are certainly at a different state from those of the anlage and thus could interfere with the pseudotime of cells from the anlage. The RNA velocity analysis using this subset strengthened our previous inferences. First, C1 seems to be the least dedifferentiated cell cluster, whose future state will be cells of C3 and part of the population of C4. Second, that C0 seems to be in a specialized state of differentiation that has a relationship to C4 (Fig 11G). This would explain the relationship of this cluster to that of the differentiating cells of C5 and C9. Interestingly, portions of C1, C0 and C4 appear to be in an advance process of differentiation based on their longer arrows compared to C3 and another portion of C4 close to C3 (Fig 11G). The Slingshot analysis of the anlage cells from C0, C1, C3 and C4 revealed that the pseudotime starts at a point of convergence that contains a portion of cells from C1 and C0. Yet, it further supports the cells from C1 as the least differentiated (Fig 11H). C1 is then followed by cells of C3, C4 and lastly C0, which for the most part seems to be at a more advanced differentiation state with a closer relationship to differentiating cells (Fig 11I).

Discussion

In this study we employed scRNA seq and HCR-FISH techniques to examine the cellular phenotypes in regenerating intestinal tissues of the sea cucumber H. glaberrima. These techniques have been seldom used to characterize echinoderm cells, and the few studies available are mainly limited to embryonic stages of sea urchins (Strongylocentrotus purpuratus, Lytechinus variegatus) and the sea stars (Patiria miniata) (Cocurullo et al., 2023; Foster et al., 2022; Meyer et al., 2023; Paganos et al., 2022a, 2022b, 2021; Tominaga et al., 2023). Nonetheless, these studies provide an excellent description of the cell population and dynamics arising from major germ lines during echinoderm development. We now apply the same techniques to explore cell phenotypes involved in intestinal regeneration in holothurians. This is, to our knowledge, the first-time scRNA-seq and HCR-FISH have been used in adult echinoderms to analyze the cellular and molecular basis of their amazing regenerative properties. This research integrates the extensive cellular and molecular information on intestinal regeneration in holothurians collected over the past two decades, offering a comprehensive view of the cellular phenotypes and molecular changes involved.

Cell types and differentiation stages in the 9-dpe regenerating intestine

Our study has focused on the description of cells present in the 9-dpe regenerating intestine of the sea cucumber, where cells are known to have gone through dedifferentiation and are, at this time-point, in the process of differentiating into specialized cells (García-Arrarás et al., 2011). Based on our analysis we have identified 13 distinct populations that form part of the regenerating intestinal mesentery and anlage (Fig 2A). One intriguing finding was the lack of significant differences in the cell clusters between the anlage and the mesentery. However, this can be explained by two different facts. First, we have previously shown that many of the cellular processes that take place in the anlage, including cell proliferation, apoptosis, dedifferentiation and ECM remodeling occur in a gradient that begins at the tip of the mesentery where the anlage forms and extends significantly into the mesentery (Garcia-Arraras et al., 2006, 2011, 2019; Mashanov et al., 2012; Reyes-Rivera et al., 2024). Similarly, migrating cells move along the connective tissue of the mesentery to the anlage (Cabrera-Serrano and Garcia-Arraras 2004).

Thus, there is no clear partition of the two regions that would account for distinct cell populations associated with the regenerative stage. Second, the two cell populations that would have been found in the mesentery but not in the regenerating anlage, mature muscle and neurons, were not dissociated by our experimental protocol as to allow for their sequencing. Current experiments are being done using single nuclei RNA sequencing to overcome this hurdle.

Among the 13 clusters we have described the major divisions are clusters corresponding to the coelomocyte, coelomic epithelium, and mesenchyme cell types (Fig 3). Thus, we can now use the information that our laboratory and others have gathered to pinpoint the cellular mediators and their activity during regeneration. The coelomic epithelia, for example, is highly influential in the regeneration process as it is the site where major cellular events occur, particularly cell division, dedifferentiation, and differentiation (García-Arrarás et al., 2011). Moreover, the mesenchyme ECM is known to undergo remodeling during the regeneration process, which is also critical for the proper growth of the new tissue (Quiñones et al., 2002). However, while much of the events have been described using microscopic and histological tools, we need a more comprehensive understanding of the transcriptomic characteristics of specific cell populations involved in the regeneration process.

The sea cucumber contains various mesenchymal and coelomocyte populations

Amongst the phenotypes identified are clusters of mesenchymal (C2 and C7) and coelomocyte cell populations (C6, C10, C11, and C12). The mesenchymal populations demonstrated a unique expression of ERG and ETS-1. ERG, in particular, has been related to embryonic development, differentiation, angiogenesis, and apoptosis (Dhordain et al., 1995; Iwamoto et al., 2001; Vlaeminck-Guillem et al., 2000) (Fig 2). More importantly, the expression of ERG has been associated with mesenchymal identity in other echinoderms (Meyer et al., 2023; Tominaga et al., 2023). In these studies, ERG has been reported as the marker gene of mesenchymal cells of sea urchin and sea star larva by scRNA-seq analyses and the localization of ERG on embryonic precursor mesenchymal cells of the sea urchin was further confirmed by in situ analyses (Meyer et al., 2023). In addition, this latter group also reported these mesenchymal cells to express a GATA transcription factor (GATA3) and ETS1, which in our dataset are also being expressed only by populations within this supra-cluster (GATA2 and ETS1 – Fig 3B). The expression of PRG4 (proteoglycan-4) reinforces the mesenchymal identity of this supra-cluster in the regenerating intestine (S5 Fig), which correlates with previous studies that have identified a proteoglycan-like molecule in the mesenchyme of 7-dpe regenerating intestine (Vázquez-Vélez et al., 2016).

ETS-1 has long been associated with developmental processes of mesenchymal formation in sea urchin (Koga et al., 2010; Rizzo et al., 2006). In other studies, ERG, a member of the ETS gene families, has been found to be necessary for controlling mesenchymal identity and differentiation (Cox et al., 2014; Mochmann et al., 2013). Further analyses of their individual marker genes suggest that these cells might be involved in EMT. For instance, C2 expresses TIMP3, a metalloproteinase inhibitor that aids in the extracellular matrix remodeling (Dewing et al., 2020), and NPNT, a gene reported to be related to development and cancer processes (Magnussen et al., 2021). These genes are also crucial for cells undergoing EMT as the cells need to detach from the other cells and the basal lamina that forms the epithelium. Comparatively, the expression profile of C7 with genes such as PA21B (phospholipase A2), TMPS9 (transmembrane protease serine 9), SEPP1, DMBT1, ITIH3, and SVPE1 and its GO results suggest this mesenchymal population is undergoing different processes. Based on these contrasting expression profiles, we propose that C2 corresponds to cells that have recently undergone EMT from the coelomic epithelium and eventually differentiate into a more specialized phenotype (C7). Our pseudotime results further support this developmental transition as cells in C7 appears to be in a more advanced stage when compared to those of C2 (S13 Fig). Therefore, these two populations correspond to the first transcriptomic description of mesenchymal phenotypes reported in sea cucumber regenerating intestine.

Our dataset contains four distinct populations that we have characterized as coelomocytes. The coelomocyte populations reported in different holothuroid species ranged between 4 to 6 distinct types (Hetzel, 1963; Ramírez-Gómez et al., 2010; Xing et al., 2008). Studies from our laboratory previously revealed four different coelomocyte populations in H. glaberrima, distinguished by their morphology, histochemistry, and phagocytic activity. These were lymphocytes, phagocytes, spherulocytes and a population named “giant cells” (Ramírez-Gómez et al., 2010). The distinctive gene expression profile of each of the coelomocytes that we have identified can provide insights into the differences in their role as immune/circulating cells within the sea cucumber. For example, the C6 marker gene fibrinogen-like protein, which is part of a protein family known as FREP, makes this population of great interest.

Mainly because these molecules have been vastly studied across invertebrates, and multiple immune roles have been proposed, including phagocyte recognition and encapsulation (Hanington and Zhang, 2011). A distinct example is that of C10, where the expression of HCK, FER and ITF8 markers suggest this might be a macrophage-like activity (Chen et al., 2023; Dolgachev et al., 2018; Shuttleworth, 2018). Specifically, HCK, a member of Src family kinases (SFK), has been closely related to macrophage activation and polarization (Bhattacharjee et al., 2011; Poh et al., 2015). Interestingly, recent studies in A. japonicus found that an Src homolog mediates the phagocytosis of Vibrio splendidus, which further supports C10 immune identity (Wan et al., 2022). Thus, to our knowledge, this would be the first report of the expression profile of distinct coelomocyte populations in an adult echinoderm species, setting up the stage for integrating these populations with the previously described ones.

The coelomic epithelia of the intestinal anlage is composed of a heterogenous population of cells

Our research findings align with previous microscopic descriptions of cells in the normal and regenerating mesothelium and coelomic epithelia. In our data, we can easily identify the cell population that forms the cluster exhibiting a proliferative phenotype (C8). Proliferative cells in the regenerating intestine are primarily localized within the coelomic epithelium of the anlage (García-Arrarás et al., 2011, 1998; Reyes-Rivera et al., 2024), and the gene expression profile documented here unequivocally identifies these as proliferative. These cell proliferation genes are specific to C8, with minimum representation in other populations. Interestingly, as mentioned before C8 expresses at lower levels many of the genes of other coelomic epithelium populations. Nevertheless, even if we mask the top 38 proliferation genes (not shown), this cluster is maintained as an independent cluster, suggesting that its identity is conferred by a complex transcriptomic profile rather than only a few proliferation-related genes. Therefore, the identity and potential role of C8 could be further described by two distinct alternatives: (1) cells of C8 could be an intermediate state between the anlage precursor cells (discussed below) and the specialized cell populations or (2) cells of C8 are the source of the anlage precursor populations from which all other populations arise. The pseudotime data is certainly complex and challenging to interpret with our current dataset, yet the RNA velocity analysis showed in Fig 11B would suggests that cells of C8 transition into the anlage precursor populations, rather than being an intermediate state. This is also supported by the Slingshot pseudotime analysis that incorporates C8 (S13 Fig). Nevertheless, additional experiments are needed to confirm this hypothesis.

A second population of cells that can be well correlated to previously described cells is that with a muscle cell phenotype (C5). The evidence suggests that this cell population represents those cells from the coelomic epithelium that are differentiating into myocytes. This evidence includes: (1) the top-expressed genes by the cells in this cluster are all muscle-associated genes; (2) the top-enriched terms are all related to muscle morphogenesis; (3) the cluster is mostly composed of cells that come from the anlage where muscle formation is known to be taking place at this stage (Murray and García-Arrarás, 2004); (4) differentiated muscle cells were not dissociated by the enzymatic procedure strongly suggesting that the muscle cells that we have identified in our data are those that are in a differentiation process, rather than fully differentiated cells closer to the body wall; (5) the cells in this cluster were identified by HCR-FISH of MYH7. These cells are localized toward the basal region of the coelomic epithelium, the region where the differentiating myocytes are known to be present (Murray and García-Arrarás, 2004).

Moreover, among the top expressed genes of this cluster is Troponin I (TNNI1), which has also been reported to be highly expressed in muscle precursor cells of the sea star embryo and immature cardiomyocytes of the chicken (Mantri et al., 2021; Tominaga et al., 2023). The high differentiation activity of these cells is also supported by our pseudotime analysis, where distinct cells within the cluster are in individual differentiation states (Fig 11D-E). Further analysis of these populations could allow us to understand the transcriptional changes these cells undergo to become fully specialized muscle cells.

An additional population in our dataset, is the neuroepithelial population (C9). This population has STARD10, among its top expressed genes, known to be a phospholipid transfer protein present in a neural cell population. This population has been localized in the coelomic epithelium of the intestine, among other sites (García-Arrarás et al., 2019; Rosado-Olivieri et al., 2017). Furthermore, cells of this cluster have high expression of other genes reported to be expressed by neuronal cells of regenerative and developmental tissues, such as beta-tubulin in developing human gut and sea urchin larva, and synapsin in planaria regenerating tissue (Cocurullo et al., 2023; Elmentaite et al., 2020; King et al., 2024). Interestingly, an earlier study showing that beta-tubulin positive cells arise from dedifferentiated cells of the regenerating intestinal tissue of the sea cucumber (García-Arrarás et al., 2019) also suggests that similar to the muscle population, these cells are neuroepithelial cells that are differentiating rather than fully specialized. Further support for this theory lies in the main contribution to C9 coming from anlage cells and their terminal differentiation state observed in the pseudotime analysis (Fig 11D).

The remaining four clusters of the coelomic epithelium supra-cluster are more challenging to characterize. Nonetheless we will explore some hypotheses regarding their cellular phenotypes. The identity of C4 was put forward based on its close transcriptional and pseudotime correlation with populations of the coelomic epithelia (Fig 3 and Fig 11). Additionally, in our in situ hybridization, its marker gene UNC5C as well as another gene SC6A5 (not shown here), are mainly expressed by the coelomic epithelial cells of the normal intestine, strongly suggesting that the cells expressing this gene in the anlage are those that will become the coelomic epithelial cells of the regenerated organ (Fig 9).

C0 remains an intriguing cell population. On one hand it appears to be closely related, by its gene expression, to other coelomic epithelial populations (Fig 3). Thus, it may represent an unknown cell population or cell stage. The in situ expression suggest that this cluster represents the cellular population of the mesentery or peritoneal coelomic epithelium. In this case we might be evidencing differences in the coelomic epithelium of the mesentery (exemplified by C0) from those of the coelomic epithelium of the intestine (exemplified by C4). Future experiments will be needed to address this controversy.

Lastly, C1 and C3 share many common genes, and represent distinct populations but are still closely associated, regardless of the clustering parameters or the statistical assessment performed. Although we considered combining them into a single cluster, we ultimately decided against it, as they likely represent different stages of cell development or plasticity in the regenerating intestine. One-to-one comparisons revealed that C1 expressed various ribosomal gene markers, while C3 expressed specific genes such as EPHA4 (ephrin type-A receptor 4) and LRIG3, indicating distinct transcriptomic states. Additionally, all trajectory analyses revealed that C1 cells appear to give rise to C3 cells.

Further characterization of C1 and C3, within the anlage coelomic epithelium, suggest that these cells probably serve as cellular precursors to differentiating cells. The evidence from pseudotime analyses and gene expression of C1 and C3 strongly supports this conclusion. These cells appear to exhibit pluripotency with the potential to form muscle, neurons, coelomic epithelia and mesenchymal cells in the regenerating intestine. Their gene expression profiles include markers from gene families associated with embryonic development such as Hox, zinc fingers, basic helix-loop-helix and others, some of which are linked to stemness and pluripotency. For example, HSP90A, a molecular chaperone essential for stem cell pluripotency, and markers like HES1 and TGFb, essential for maintaining stem cell proliferation, are present in these cell populations (Aztekin, 2021; Mishra et al., 2005).

Moreover, the expression of PIWL1 (piwi-like protein 1) and YAP1 further support the classification of these clusters as precursor cells of the intestinal anlage (S12 Fig). In hydra, piwi-like molecules are exclusively localized in stem/progenitor cells (Juliano et al., 2014), and in mice, YAP1 is expressed only in multipotent cells during intestinal epithelium regeneration, vital for their emergence (Ayyaz et al., 2019). Ultimately, trajectory analysis also supports C1 as a precursor cell population, as the dividing cells (C8) appear to give rise to C1, which then progresses toward C3. This analysis aligns with evidence of a proliferation center in the coelomic epithelia that provides precursor cells essential for the growth of the anlage.

In summary, C1 and C3 cells are probably dedifferentiated cells from the mesentery mesothelium, retaining markers common to all epithelial cells. Similar to what is known in other highly regenerative organisms, these dedifferentiated cells can proliferate and then re-differentiate into the cells of the new organ. Future experiments will determine whether they retain the memory of their previous phenotype (muscle or coelomic epithelium) or are completely pluripotent.

The coelomic epithelium of the intestinal anlage is pluripotent

Our study reveals that the coelomic epithelium, as a tissue layer, is pluripotent. It fulfills the definition of pluripotency; a group of cells that has the ability to differentiate into many, but not all, cell types. Microscopy studies across different echinoderm species have consistently suggested that the coelomic epithelium can differentiate into various cell types. Among these are the formation of muscle cells and neurons in sea cucumbers (Yu Dolmatov et al., 1996), mesenchymal cells in brittle star (Piovani et al., 2021), and even immune cells (coelomocytes) in sea star (Byrne et al., 2020; Sharlaimova et al., 2021). In some species, the coelomic epithelium has even been proposed to transdifferentiate into intestinal luminal cells (Mashanov et al., 2005). In our analysis we can identify the cells in the coelomic epithelium that are differentiating toward muscle (C5), neurons (C9), coelomic epithelium (C4) and mesenchymal cells (C2 & C7) corroborating what has been previously described in H. glaberrima studies. Nonetheless, it remains unclear whether individual cells are pluripotent, thus a more definitive conclusion will be reached once lineage tracing experiments can be done.

The coelomic epithelium of the regeneration anlage could also be the source of other cell types. This is exemplified in the brittle star Marthasterias glacialis, where it has been suggested that the coelomic epithelium is not only involved in arm regeneration but also serves as a source of immune cells (Guatelli et al., 2022). This report indicated that “residential stem cells” in the regenerating arm of the brittle star originate from the coelomic epithelia (Candia-Carnevali et al., 2009). Similarly, in the sea cucumber Holothuria forskali, the injured mesothelial layer has been identified as the source of undifferentiated cells that differentiate into the cells of the growing organ and into phagocytic cells, now recognized as coelomocytes (Vandenspiegel et al., 2000).

Of particular interest, our study reveals that the mesentery coelomic epithelia population (C0) exhibits localized expression of SAA1, an immune response related gene. This finding, in line with previous reports of SAA1 localization in the coelomic epithelium of the regenerating intestinal tissue (Santiago et al., 2000) (S12 Fig), raises the intriguing possibility that the coelomic epithelium of the sea cucumber regenerating intestine could also be the source of coelomocytes, as suggested by Guatelli (2022) (Guatelli et al., 2022). This hypothesis is further supported by the observation that in the sea star Asteria rubens, coelomocytes arise from its coelomic epithelium (Vanden Bossche and Jangoux, 1976). However, our current results, while suggestive, do not yet provide conclusive evidence to support this in the sea cucumber, at least not in the coelomic epithelia of the 9-dpe regenerating intestine.

Currently, there is no definitive evidence identifying the precursor cells that give rise to these cell populations in echinoderms. However, given our gene expression data and the interactions observed among the cell populations, we postulate that cells from C1 stand as the precursor cell population from which most of the cells in the anlage coelomic epithelium arise. This is based on their preferential localization within the anlage coelomic epithelium, their lack of any clear differentiated cell marker, and their association with the cell dividing cluster. Granted, much more experimental evidence will be needed to arrive at this conclusion. Nonetheless, it provides a focus on the most likely, and probably most intriguing cell population. It also provides the path to explore multiple questions that arise from our results. Can C1 be subdivided into other cell populations? Are the cells in this cluster pluripotent? Do these cells originate from the same cells via dedifferentiation? These and many other questions will indeed be tested in future experiments.

The intestinal anlage as a regeneration blastema

The intestinal anlage is a particular regenerative structure that has long been described as a blastema-like structure due to its remarkable morphological resemblance to the “classical blastema” (García-Arrarás et al., 1998). As stated previously a blastema is usually described as a transient structure composed of a mass of proliferating undifferentiated cells. This structure is found at the site of injury and will give rise to the regenerated organ (Seifert and Muneoka, 2018). The blastema cells can originate from different lineages across species. For example, in some amphibians the blastemal cells originate from dedifferentiated muscle, cartilage, fibroblast and other tissues, while in others they include cells that originate from muscle satellite cells, a type of stem cell (Globus et al., 1980; Sandoval-Guzman et al., 2014). In Planaria they originate from undifferentiated stem cells known as neoblasts (Baguñà, 2012; Wagner et al., 2011). The blastema is overlayed by a wound epidermis that is formed by a re-epithelization process following injury, which develops into a specialized wound epidermis that in amphibians is known as the apical epithelial cap (AEC) (Aztekin, 2021). Since its first description, more than 100 years ago, the blastema has been regarded as the best indicator of regeneration, and its presence is associated with tissues or organs that are highly regenerative. In fact, it has been proposed that the presence or absence of a blastema defines the regenerative success or failure of the regeneration process. However, as more regenerative species continue to be studied, the classical definition of a blastema has been reconsidered, moving towards its functional role rather than its histological or structural characteristics (Seifert and Muneoka, 2018).

The blastema of salamanders and newts, have long served as models to describe a blastema cellular and molecular properties (Globus et al., 1980; Scimone et al., 2022; Tajer et al., 2023). The sea cucumber intestinal anlage stands as a different structure in terms of the tissue compartmentalization of certain activities. Yet, when examined closely, much of the processes and signaling molecules described in the amphibian blastema also take place in the holothurian anlage, although their spatial occurrence might differ. For instance, we and others have previously described the dedifferentiation process by which muscle cells dedifferentiate and proliferate (Garcia-Arraras et al. 2011; Dolmatov 2021). Thus, similar to some amphibian blastema cells, the cells of the holothurian anlage coelomic epithelium are proliferative undifferentiated cells that originated via a dedifferentiation process. Moreover, some of the genes expressed by the amphibian blastema cells or the overlying AEC, such as Wnt, Tgf-beta, and Fgf, are also known to be expressed during sea cucumber regeneration (Auger et al., 2023; Zeng et al., 2023). In amphibians, some of these factors are thought to be released by cells of the AEC, serving as a way of modulating the blastema cell activity. For example, FGF1 is expressed in the cells of the AEC while FGF receptors were localized to blastema cells (Zenjari et al., 1996). Moreover, both FGF and Wnt signaling have been shown to be vital for the formation of the blastema (Makanae et al., 2014).

This expression of Fgf and Wnt and their possible functions correlate with what has been found in sea cucumbers. Not only are the same growth factors expressed in C1 and C3, but a recent study highlighted the role of FGF4 as a modulator of cell proliferation during the intestinal regeneration of the sea cucumber A. japonicus (Zeng et al., 2023). This group demonstrated that inhibiting FGF4 and its receptor, FGFR2, negatively affected cell proliferation of the mesothelial layer during intestinal regeneration. Similar results have been shown for Wnt, primarily a study by our laboratory demonstrating that Wnt pathway inhibition, either by pharmacological drugs or by RNAi caused a reduction in cell proliferation (Alicea-Delgado and García-Arrarás, 2021; Bello et al., 2020). Thus, in the sea cucumber coelomic epithelium of the anlage, cells are also involved in intercellular communications that modulate cellular dynamics through factors similar to those found in the AEC-blastema cell modulation.

The holothurian anlage and the amphibian blastema exhibit some differences. In the latter there is a clear separation of the AEC and the underlying blastema cells. In amphibians, the AEC has been shown to actively participate in the formation and maintenance of the blastemal cells, particularly in their proliferation and differentiation. In contrast, in holothurians, cells with both blastema and AEC characteristics can be found within the coelomic epithelium. In fact, both PRRX1 (paired mesoderm homeobox protein), a gene expressed by amphibian blastema cell precursors (Gerber et al., 2018; Lin et al., 2021), and HES1 (transcription factor HES-1), a marker gene of the amphibian AEC, are expressed by the holothurian coelomic epithelium (Aztekin, 2021) (Fig 3 and S12 Fig).

Our data has allowed us to construct a model of the cell populations we identified in the 9-dpe intestinal anlage (Fig 12). This model presents the coelomic epithelium of the anlage as a heterogeneous layer of cells comprising of six cell populations corresponding to distinct differentiation states. Among these are cell populations that seem to be in the process of differentiation toward muscle (C5), neuroepithelium (C9), and intestine coelomic epithelial cells (C4). We also propose the presence of undifferentiated (C0, C1, C3) and proliferating cell populations (C8) in the coelomic epithelia, which give rise to the cells in this layer. Underneath the coelomic epithelia are the mesenchymal cells (C2, C7), some of which may have originated from the coelomic epithelial layer via EMT. Moreover, the sea cucumber intestinal anlage behaves as a blastema since it uses similar mechanisms to fulfill the same role in the regeneration of the new organ. In this case, the cells of the coelomic epithelium would be those considered akin to the blastemal cells. Our results, thus provide an alternative view of cell precursors and regenerating structures with similarities and differences to those of better studied regenerating animal models.

Model of cellular organization of the intestinal anlage.

Various cell populations can be identified in the connective tissue (CT) and coelomic epithelium (CE) in the anlage of the sea cucumber H. glaberrima at 9 dpe. These populations correspond to those identified by the scRNA-seq that are described in the text. Most of the cell populations are found within the CE, some of them showing particular localizations; differentiating muscle cells are found in the basal part of the CE while differentiating coelomic epithelia are found in the apical region. The two mesenchymal populations are shown together. The colors of cells and cluster numbers (C#s) correlate with those used on the UMAP shown in Fig 2. Inset provides the approximate localization of the diagram cells.

Methods

Animals and Sample Collection

Two adult sea cucumbers were collected from the northeastern rocky region of Puerto Rico and were kept in aerated sea water for acclimatation prior to experiments. Evisceration of the sea cucumber was stimulated by intracoelomic injection of 0.35 M KCl as previously described (Reyes-Rivera et al., 2024). They were maintained in sea water aquaria for 9 days to undergo regeneration until dissection. Before the dissection, animals were anesthetized by immersion in ice-cold water for 45 minutes. Dissection was done through an initial dorsal incision which allowed exposition of the internal organs, upon which the growing anlage (or rudiment) and mesentery were dissected and separated from each other. Each tissue was kept in CMFSS on ice and treated separately during the tissue dissociation process. Thus, tissues were dissected for four scRNA-seqs; two anlage samples and two mesentery samples.

Tissue Dissociation

Mesentery and anlage tissues were digested for 15 minutes, in rocking shaker at room temperature with 1mL of 0.05% Trypsin/0.02% EDTA solution prepared in CMFSS (Bello et al., 2015). Digestions were quenched by adding 500µL of 0.2% BSA in CMFSS and then centrifuged for 2 minutes at 1500 rpm. The supernatants were discarded, and the pellets resuspended with 500µL of 0.04% BSA in CMFSS and the cells were gently separated by pipetting using glass pipets with fire blunted tips. The cell suspensions were filtered using a nylon cell strainer with 70mm mesh and aliquots were taken for cell counting. Cell viability was assessed through manual cell counting with hemocytometer to confirm that a viability higher than 90% was maintained. Sample suspensions were adjusted to 1000 cells/mL using CMFSS, as required for sequencing procedures.

Immuno- and cytochemistry

Fifty µL of dissociated cell sample (from the same samples that were used for scRNAseq) were placed on polysine-treated slides and fixed with 50 µL of 4% paraformaldehyde and left to dry overnight. Slides were washed in PBS prior to use for immune and or cytochemistry. The methodology for the immunofluorescence techniques/preparation of slides was followed as published except for the time of PBS washes which were of 10 minutes instead of 15 minutes (Diaz-Balzac et al., 2007; Reyes-Rivera et al., 2024). The primary and secondary antibodies used are in S14 Table. Samples to be probed with fluorescent phalloidin were treated directly with Phalloidin-TRITC (1:1000; Sigma P1951) for 1h as described previously (Reyes-Rivera et al., 2024). DAPI was incorporated into the mounting medium as described previously (Reyes-Rivera et al., 2024). Cells were observed in a Nikon DS-Qi2 fluorescent microscope.

Single Cell RNA Sequencing and Data Analysis

Single cell libraries were prepared using the 10X Genomics Chromium Next GEM Single Cell 3’ Kit v3.1 and the Chromium 10X instrument, following the protocol from manufacturer (CG000204). Sequencing was carried out with Illumina NextSeq 2000 at the Sequencing and Genotyping Facility of the University of Puerto Rico Molecular Science Building.

The raw sequencing reads from all four samples (two mesentery and two anlage) were processed individually using Cell Ranger (v7.1.0) (Zheng et al., 2017) using the reference genome and gene models of H. glaberrima available at blastkit.hpcf.upr.edu/hglaberrima-v1. Gene models were annotated against human reference proteins from Uniprot and the entire Uniprot reference protein database. Gene IDs throughout the study correspond to the annotations with the human reference protein sequences from Unitprot.

Samples quality was assessed through the number of UMI, genes and overall complexity of gene expression. The number of UMI and genes was found to be over 500 and 300, respectively (Figure S4). Furthermore, the distribution of genes per cells was found to be unimodal. In general aspects, samples from each animal were similar, with the total number of cells from the anlage tissue was higher than that of the mesentery. General sequencing statistics after mapping reads with Cell Ranger can be found in S15 Table.

Initial quality filtering was conducted in R (v4.3.2) using SoupX (v1.6.2) with default parameters to remove ambient RNA (Young and Behjati, 2020). Subsequent data processing was carried out using Seurat (v4.3.0.1) (Hao et al., 2024, 2021; Satija et al., 2015; Stuart et al., 2019). After data normalization and identifying variable features (n=2000), 22,970 anchors were identified using 20 dimensions. The datasets were integrated using the canonical correlation analysis (CCA) approach of Seurat, followed by data scaling for principal component analysis (PCA) (npcs = 50) and Uniform Manifold Approximation and Projection (UMAP) dimensional reduction technique (dims = 1:20).

Neighbors were then identified with 20 dimensions, leading to cluster identification at a resolution of 0.5. Various resolution parameters (0.8, 1.2, and 1.4) and dimensions (up to 35 in increments of 5) were tested before the final resolution value. The statistical analysis tool scSHC was employed to assess the probability of each cluster being unique (Grabski et al., 2023). Marker genes of each cluster or supra-clusters were identified using the FindMarkers() function in Seurat and mapped to the appropriate Uniprot IDs. Genes highlighted throughout the study where further validated using the NCBI non-redundant reference database and EchinoBase (Telmer et al., 2024) via BLAST.

Differential expression data across clusters was used to perform gene set enrichment analysis of Gene Ontology (gseGO) biological processes terms using clusterProfiler (v.10.0) (Yu et al., 2012) with a p-value cutoff of 0.05. For this analysis, BLASTp was used to map H. glaberrima gene models to human reference protein sequences from NCBI (GCF_000001405.40), facilitating the assignment of ENTREZ ID to the correct human GO terms in the AnnotationDbi (v1.64.1) human database (v2.1) (Pagès et al., 2024).

RNA velocity loom files were generated with velocyto (v0.17), which relied on genome-masked regions obtained from RepeatModeler (v2.0.5) and RepeatMasker (v4.1.5) (Smit et al., 2015; Smit and Hubley, 2015), along with the sea cucumber gene models and CellRanger dataset. These files were further analyzed with velocyto.R (v0.6) and SeuratWrappers (v0.2.0), using the ReadVelocity and RunVelocity functions. Pseudotime analysis for the distinct data subsets was conducted using Slingshot (v2.10.0) (Street et al., 2018). The code of the data analysis has been made available at GitHub (https://github.com/devneurolab/scRNAseq_Hglaberrima).

HCR-FISH

Regenerating intestines from animals eviscerated 8-9 days previously and intestines from non-eviscerated (controls) animals were collected and fixed in 4% (v/v) paraformaldehyde with phosphate-buffered saline (0.01M PBS; 0.138M NaCl; 0.0027M KCl; pH7.4) overnight at 4 ℃. Tissues were then, washed three times by PBS, and treated overnight with 40% saccharose prior to cutting in the cryostat. Sections (20 mm) were prepared in a cryostat (Leica CM1850), as previously published for immunohistochemistry (Reyes-Rivera et al., 2024). Gene spatial expression was determined by designing twelve (12) or twenty (24) set split probes for each gene marker. To decrease probe unspecific binding and increase its signal to background ratio, probes were designed as described by H. Choi and colleagues (Molecular Instruments) (Choi et al., 2018). To ensure probe gene target specificity, all nucleotide regions selected for probe design underwent extensive validation using the newly developed H. glaberrima genome and transcriptome alignment tool (https://blastkit.hpcf.upr.edu/hglaberrima-v1/) (Medina-Feliciano et al., 2021). Probes were used at a final concentration of approximately 20-25 nM.

Immediately after slide preparation, HCR-FISH v3 was carried out using a modified version of Molecular Instrument’s fresh fixed frozen tissues protocol. Probe hybridization was conducted overnight at 37 ℃, while DNA hairpin amplification (B1-546nm or B2-546nm) was done at 25℃ overnight with 3pmol of h1 and h2, respectively. Our protocol modification involved replacing Ethanol with Methanol during sample permeabilization. Also, the use of proteinase K was omitted, and Tween 20 at 0.1% was added to all PBS wash buffers. Once slides were prepared, image acquisition was attained using a Nikon DS-Qi2 fluorescent microscope. Positive control probes (PolyA) were used as signal calibrators to define background from positive signal during image analysis (S10 Fig). Negative controls with fluorescent hairpins only can be found in S11 Fig. Each probe was studied in at least 3 different animals, none of which were the specimens used for the scRNA-seq.